\chapter{Business}

\epigraph{Risk comes from not knowing what you're doing.}{Warren Buffett}

How to succeed in business.

TLDR:

1. Find something you unconditionally love to do.
2. Find how to find buyers.
3. Find a way to run the business without yourself.
4. Validate your market.
5. Carry out the plan.

1. What is something you unconditionally love to do?
2. Are you sure that you are quite strong/competitive on that?
3. Why will they buy from you?
4. How will they find you?
5. Can you make them your fanatics? (Fans are irrational and we want our buyers to be irrationally loyal.)
6. How can that business run without yourself?

There are only two prerequisites to business:

1. Find a thing that you will still do even if you don't get any money from it.
Find a thing that you will still try to do even if you are on your deathbed.
Find a thing that you are internally motivated to do.

2. Find a way to get people to pay you for that.

3. Validate your market.

Step 1 is important because you must hang on, and the only thing
that keeps you from giving up despite all hardship is your internal motivation.
Success requires persistence.
If you are internally motivated, you will do it without money.
If you are not, money won't make you do it.

Step 2 is important because if you don't work your own land,
you will need money to survive.
Doing what you love is not enough. You must find a buyer.

There is no need for KPI, metrics, whatnot.
Just make sure that your income exceeds your expense.
Things will roll from there.

Why people buy things:

1. They already know what they want.

2. They don't know what they want.

How do people know what to buy?
They know it when they see it.

Somehow some people get interested in some things.
The best way to get your things sold is to let as many people as possible know.

Fans. These people always look out for things they love.
Start with the fans.

Followers. Friends of fans. Buy because friends buy.
