\chapter{Notation}

$a : A$ or $a \in A$ means the type of $a$ is $A$,
or $a$ is in $A$,
or $a$ is an element of $A$.

$A \to B$ is the type of every function
that takes any $a \in A$ and returns any $b \in B$.

Concatenation means function application:
$fx$ means $f(x)$ in standard mathematical notation.
$fxy$ means $f(x,y)$ in standard mathematical notation,
but the former one is with currying.
$f0$ can mean $f(0,y)$ which is a function of $y$ only.

If $f : A\to B$ and $x:A$, then $fx : B$.

If $f : A\to B\to C$, $x:A$, and $y:B$, then $fxy : C$.





\begin{mcor}[number of deciders]
    Let $|X| = m$.
    The number of total mappings of type $X^n \to \mathbb B$ is
    \begin{align}
        2^{m^n}
    \end{align}
    The number of total mappings of type $(X^0 \cup X^1 \cup \ldots \cup X^n) \to \mathbb B$ is:
    \begin{align}
        \sum_{n=0}^{\infty} 2^{m^n}
    \end{align}
    There are $m^n$ boxes and each box can contain either $\syes$ or $\sno$.
\end{mcor}





\begin{mdef}[Ambiguous operator]
    The \emph{ambiguous operator} $\amb : A^* \to A$
    takes a list and returns an element of that list
    such that the decider containing the amb-expression is led to $\syes$.
\end{mdef}

    The following is a nondeterministic algorithm:
    \begin{align}
        \SDP fx &= gfx[] & \text{where}
        \\
        gfxy &= \fite{\fnull x}{fy}{gf(\ftail x)(\amb[y, \ftail y])}
    \end{align}






\begin{mlem}[existence of decider bounded below by LIN]
There exists a decider that cannot be computed
in less than linear time
by the standard machine.
Formally there exists an $X^*$-decider $f$ such that
for each list $x : X^*$ of length $n$,
$\TC(fx) \in \Theta n$.
\begin{proof}
    Consider the decider that decides whether zero
    is the modulo-$m$ sum of all elements in the input list of $X$s
    where $X$ is a finite subset of $\mathbb N$ of size $m$.

    $f x = [\sum_{k=0}^{n-1} x_k = 0]$
\end{proof}
\end{mlem}

    \begin{align}
        \ambc y &= \fite{\fnull y}{y}{\amb[\fhead y :: \ambc (\ftail y), \ambc (\ftail y)]}
    \end{align}

\begin{mcon}
Corollary:
If $\TC_D(\SDP fx) \in \Omega(2^n \cdot n)$ and $\TC_N(\SDP fx) \in O n$,
there exists a problem that is in NP but not in P;
therefore $P \subset NP$.
\end{mcon}

Prove that the only $g$ that satisfies $fx = f(gx)$ is $\text{id}$.

To prove that $\TC (fx) \in O(gn)$,
write an algorithm that is in that time complexity class.

To prove that $\TC (fx) \in \Omega(gn)$, ...


Function Constancy Check Problem/Function Constancy Decider.

Given an almost constant $X^n$-decider, find the only $x$ that differ.

An $X$-decider is almost constant iff $f^{-1}\syes = \{\}$.

This will also prove P neq NP:
Find $f$ and $g$ such that $\forall x \in X^* ~ fx = gx$ but $\TC_D(fx) \subset \TC_N(gx)$









\chapter{Computational complexity}

A set is \emph{computable} iff there is an algorithm (a finite description)
for its indicator function.

\section{Lax monoidal functors}

A category $C$ is defined by $\ob C : \Set\alpha$ and $\hom C : \Set(\alpha, \alpha))$ such that
associative and preserve identity
\begin{enumerate}
    \item $(f \circ g) \circ h = f \circ (g \circ h)$ for all $f,g,h : \hom C$.
    \item $1_C \circ f = f \circ 1_C = f$ for all $f : \hom C$.
\end{enumerate}

Functor is $F$:
\begin{enumerate}
    \item $f : (A \to B) \to (FA \to FB)$.
\end{enumerate}

Also known as `monad' in Haskell.

A lax monoidal functor $F$ takes a set.
\begin{enumerate}
    \item Zero element $F_0 : FA$.
    \item Lifting morphism $F_1 : A \to FA$.
    \item Internal operation on Kleisli category $f : FA \to (A \to FB) \to FB$.
\end{enumerate}

\section{Monads}

Let there be the following type functions:
\begin{align}
    \mM A &= \{ \mM_0 \} \cup \{ \mM_1 a ~|~ a \in A \}
 \\ \mL A &= \{ \mL_0 \} \cup \{ \mL_1 h t ~|~ h \in A, ~ t \in \mL A \}
 \\ \mP A &= \{ \mP X ~|~ X \subseteq A \}
\\ \mE AB &= \{ \mE_0 a ~|~ a \in A \} \cup \{ \mE_1 b ~|~ b \in B \}
\end{align}
In Haskell, $\mM$, $\mM_0$, and $\mM_1$ are known as Maybe, Nothing, and Just;
and $\mL$, $\mL_0$, and $\mL_1$ are known as list type function ([]), the empty list ([]), and the list constructor (:).

\begin{align}
    \sigma_M &: MA \to (A \to MB) \to MB
\end{align}

For \mM, $\sigma$ is defined as
\begin{align}
    \sigma \mM_0 f &= \mM_0
    \\ \sigma (\mM_1 a) f &= fa
\end{align}

For \mL, $\sigma$ is defined as
\begin{align}
    \sigma \mL_0 f &= \mL_0
    \\ \sigma (\mL_1 h t) f &= [\amb (fh \cdot \sigma tf)]
\end{align}
where $\amb : \mL A \to A$ is McCarthy's ambiguous operator
and $\cdot$ is list concatenation.
The operator $\amb$ picks an element leading to accepting path
in zero time without computing the other elements.

For \mP, $\sigma$ is defined as
\begin{align}
    \sigma (\mP S) f &= \{ \amb \{ f x ~|~ x \in S \} \}
\end{align}

\section{\mM-extended string}

\begin{align*}
    x_n' &= \begin{cases}
 \mM_1 x_n &: n \text{ is a valid index into input string}
  \\ \mM_0 &: \text{otherwise}
\end{cases}
\end{align*}

\section{Definition of computation via expression reduction}

The grammar for expression:
\begin{align}
    Exp &::= X
      \\ &| -Exp
      \\ &| (Exp)
      \\ &| Exp + Exp
      \\ &| Exp \cdot Exp
\end{align}

A finite countable set $X$.

There are $n^{n \cdot n}$ possible binary operations on a set of $n$ elements.
We assume using space-time-tradeoff that some of them can be performed in constant time.

A \emph{primitive expression} is an expression whose time cost is one unit.
We assume that these are the \emph{only} primitive operations:
\begin{enumerate}
    \item modulo-$m$ arithmetic operations:
            negation ($-a$),
            addition ($a + b$),
            subtraction ($a - b$),
            and multiplication ($a \cdot b$).
    \item every unary operation in $N_m$ and every binary operation in $N_m$
    \item every unary operation in $\mathbb B = \{0,1\}$ and every binary operation in $\mathbb B$
    \item comparison: comparing equality $(=) : X \to X \to \mathbb B$
        and (unsigned) ordering $(<) : X \to X \to \mathbb B$,
        and three-way conditional ($Cctf$: if $c$ then $t$ else $f$ where $t,f:\alpha$ and $c:\mathbb B$).
    \item primitive list operations:
        \begin{enumerate}
            \item \fnull: determining whether a list is empty
            \item \fhead: taking the first element (partial function)
            \item \ftail: taking all but the first element (partial function)
            \item constructing a list by prepending one element to a list (::)
        \end{enumerate}
\end{enumerate}
For each constant expression $e$, $\tau e = 0$.

A \emph{primitive value} is a value whose space cost is one unit.
\begin{align}
    \forall x \in X \quad \sigma x &= 1
    \\ \forall x \in (A,B) \quad \sigma x &= \sigma a + \sigma b
     \\ \forall fx \in \{L a | a \in A\} \cup \{R b | b \in B\} \quad \sigma (fx) &= 1 + \sigma x
\end{align}
Example:
\begin{align}
    \sigma[1,2,3] &= \sigma(1::2::3::[])
               \\ &= \sigma (L1(L2(L3L_0)))
               \\ &= 1 + \sigma 1 + \sigma (L2(L3L_0))
               \\ &= 1 + 1 + 1 + \sigma 2 + \sigma (L3L_0)
               \\ &= 3 + 1 + 1 + \sigma 3 + \sigma L_0
               \\ &= 5 + 1 + 1
               \\ &= 7
\end{align}

For example, the following $f$ performs modulo-$m$ summation:
\begin{align}
     fx &= C (\fnull x) 0 (\fhead x + f(\ftail x))
\end{align}

The space and time cost for reducing an expression is $\sigma E$ and $\tau E$ respectively.

\section{Definitions}

A \emph{problem} is a set of its instances.
An instance is a pair of question and answer.
The question is a string of the input alphabet.
The answer is a string of the output alphabet.
A question may have more than one answer.
Each question must have at least one answer.

An \emph{$\alpha$-machine} has a \emph{transition function}
$t : (S,\mM X) \to \alpha S$
where $S$ is the \emph{state type},
$X$ is the \emph{input alphabet},
and $\alpha$ determines the way the machine performs computation.
$X$ must be finite and countable.
$S$ must be countable.
$S$ may contain a special elements:
the accepting state \syes.
Formally, a \emph{machine} is a 4-tuple $(\alpha,X,S,t)$ where
\begin{align}
    X &: \Fin
 \\ S &: \Set
 \\ t &: (S,\mM X) \to \alpha S
\end{align}

Once something is accepted or rejected, it is final:
\begin{align}
    \forall x \quad t (\syes,x) &= \pi \syes
  \\ \forall x \quad t (\sno,x) &= \pi \sno
\end{align}
where $\pi : S \to \alpha S$ is the $\alpha$-lifting function.

The transition function must be total
and it must be an semi-algorithm;
it must be able to be described
in a finite number of (what?) elementary operations.
It must go into an infinite loop when the answer is no.

A \emph{path} of that machine for input $x = [x_0,\ldots] \in X^*$ is $s = [s_0,\ldots] \in S^*$.
The computation performed by that machine is a repeated application of the transition function:
\begin{align*}
    s_n &= \sigma (t (s_{n-1}, x_n'))
\end{align*}
where $\sigma$ is defined according to $\alpha$
and $x'$ is \mM-extended $x$.
The initial state is $s_0 \in S$.
The path is an \emph{accepting path} iff $\syes$ is in that path.
The machine \emph{accepts} an input iff the path for that input is an accepting path.

The \emph{running time} for that input is the length of the corresponding path.

The \emph{maximum space usage} is: (maximum or supremum?)
\begin{align}
    \max_{x \in s} (\mu x)
\end{align}

Deterministic transition function type is $(S,\mM X) \to S$.
Nondeterministic transition function type is $(S,\mM X) \to \mP S$.

Nondeterministic machine can \emph{look into the future}
to \emph{guess an accepting path};
deterministic machine cannot.
Given the same input, a deterministic machine
will always go through the same path of computation.
Given the same input, a nondeterministic machine
might go through another path of computation.


We say that a language is in $\aTIME f$ iff
there exists a constant $k$ and an $\alpha$-machine
such that the length of the shortest path of that machine
for each acceptable input $x$ is in $O(f(\mu x))$.

\begin{align}
    \DTIME &= \idTIME
 \\ \NTIME &= \mPTIME
\end{align}

\section{Example machines}

This machine accepts all inputs:
\begin{align*}
    S &= \{ \syes \}
    \\ s_0 &= \syes
 \\ t(s,x) &= \pi s
\end{align*}
This machine does not accept any input:
\begin{align*}
    S &= \{ \sno \}
    \\ s_0 &= \sno
 \\ t(s,x) &= \pi s
\end{align*}
This machine accepts the language $\{1\}^*$:
\begin{align*}
    S &= \{ s_0, \syes, \sno \}
    \\ t(s,x) &= \text{case $(s,x)$ of} \begin{cases}
      (s_0, \mM_1 1) &\pi s_0
        \\ (s_0, \mM_0) &\pi \syes
 \\ \text{otherwise} &\pi \sno
\end{cases}
\end{align*}

Assume that $X = \{0,1\}$.
\begin{enumerate}
    \item How many languages can an $\alpha$-machine with $n$ states accept?
    \item How many states are needed for an $\alpha$-machine to recognize a given language?
    \item How many transition functions are possible for a machine with $n$ state?
\end{enumerate}

\section{Languages accepted by a machine}

Let $\mu X = \mu Y = 2$.

Types are measurable:
\begin{align}
    \mu(A\to B) &= \mu A \cdot \mu B
    \\ \mu(A,B) &= \mu A \cdot \mu B
    \\ \mu(\mM A) &= 1 + \mu A
    \\ \mu(\mL A) &= 1 + \mu A \cdot \mu(\mL A)
    \\ \mu(\mP A) &= \mu(2^A) & \text{if $A$ is countable}
\end{align}
Hypernatural measures:
\begin{align}
    \mu A = \omega &\implies \mu(\mM A) = \omega
    \\ \mu A > 0 &\implies \mu(\mL A) = \omega
    \\ \mu A < \omega &\implies \mu(\mP A) = 2^{\mu A}
\end{align}

How many transition functions does a \mM-machine have?

How many languages can be accepted by \mM-machine with $\mu S = 2$?

Suppose that a problem is in NP.
Then there exists a constant $k$ such that
the length of the shortest nondeterministic path
for valid input whose size is in $O(n)$ is in $O(n^k)$.

The time complexity measure of a computation is
the length of the path of that computation.

The set of regular languages is $\DSPACE(O(n)) \cap \DTIME(O(n))$.

\section{Speedup theorem?}

Alphabet size can be exchanged with state size.
Space can be traded for time?

\section{Example: modulo-$m$ subset-sum problem}

$X = \{0,1,2,3,\ldots,m-1\}$.
$\syes$ iff there exists $N \subseteq \mathbb N$ such that
\begin{align}
    \sum_{k \in N} x_k &= 0
\end{align}
where addition is done modulo-$m$.

The modulo-$m$ subset-sum problem is in $\NTIME(O(n))$ and is in $\NSPACE(O(m))$
where $n$ is the length of the input.

\begin{align}
    S &= \{ S_0 \} \cup \{ S_1 n ~|~ n \in X \}
    \\ s_0 &= S_0
\end{align}
$t$:
\begin{align}
    (S_1 0, x) &\to \{ \syes \}
    \\ (S_1 s, \mM_1 x) &\to \{ S_1 s, S_1 (s+x) \}
    \\ (S_0, \mM_1 x) &\to \{ S_0, S_1 x \}
\end{align}

\section{Some Haskell code}

\begin{verbatim}
type Tr s x y = (s, Maybe x) -> (Maybe s, Maybe y)

run :: Tr s x y -> s -> [x] -> [y]
run tr state input =
    let
        run' s xs =
            let
                (mx,xs') = case xs of
                    [] -> (Nothing, [])
                    x : t -> (Just x, t)
                (ms',my) = tr (s,mx)
                ys = case ms' of
                    Just s' -> run' s' xs'
                    Nothing -> []
            in
                maybe id (:) my ys
    in
        run' state input
\end{verbatim}

\section{Example: 3-SAT in NP}

Encode a boolean expression.

$X = \{ T, F, A, 0, 1, 2, N, C, D \}$

The grammar for the input is \verb@inp@:
\begin{verbatim}
inp     ::=     asgs A exp
asgs    ::=     empty | asg asgs
asg     ::=     T | F
exp     ::=     0 | 1 | 2
        |       N exp
        |       C exp exp
        |       D exp exp
\end{verbatim}

$A$ separates the assignment and the expression.

The state type is a tuple of three boolean values.

For example, the expression $v_0 \wedge v_1 \vee \neg v_2$ with truth assignment $(v_0,v_1,v_2) = (T,F,F)$ is encoded as TFFDC01N2.

There are $n$ variables.

SAT is in NP iff SAT is verifiable in P.

Constant space. Linear time.

\section{Is there a bijection?}

\begin{align}
    \mathbb N &\iff 2^\mathbb N
    \\ 0 &\iff \{\}
    \\ 1 &\iff \{0\}
    \\ 2 &\iff \{1\}
    \\ 3 &\iff \{0,1\}
    \\ 4 &\iff \{2\}
    \\ 5 &\iff \{0,2\}
    \\ 6 &\iff \{1,2\}
    \\ 7 &\iff \{0,1,2\}
\end{align}
will enumerate all of $2^\mathbb N$?

But Cantor's theorem.

We can construct $x \in 2^\mathbb N$ that has no counterpart in the above bijection.
Counterintuition of infinity.

Binary numbering of $2^\mathbb N$.

