\chapter{Matrices}

The plural of \emph{matrix} is \emph{matrices}.

Matrices were invented to abbreviate systems of linear equations?%
\footnote{\url{https://en.wikipedia.org/wiki/Matrix_(mathematics)\#History}}

A matrix is a rectangular array of numbers.

We write \(A_{ij}\) or \(A(i,j)\) to mean the cell at row \(i\) column \(j\).

\paragraph{Finding the dimension of a matrix}

We say that the dimension of a matrix \(A\) is \((m,n)\),
written \(\dim(A) = (m,n)\),
iff the matrix has \(m\) rows and \(n\) columns.

\paragraph{Example}
The following matrix has 2 rows and 3 columns, so its dimension is \((2,3)\).
\Formula{\NoNumber
    \Matrix{
        A_{11} & A_{12} & A_{13}
        \\
        A_{21} & A_{22} & A_{23}
    }
}

\section*{Don't confuse vectors and coordinate tuples}

Use the phrases \emph{column matrix} and \emph{row matrix}
instead of \emph{column vector} and \emph{row vector}.%
\footnote{\url{https://en.wikipedia.org/wiki/Row_and_column_vectors}}

The phrase \emph{column vector} confuses a geometric vector and
the coordinate tuple that describes the vector.

A vector is \emph{not} a bunch numbers.
A vector is a geometric object with length and direction.
A vector can be \emph{described} by a coordinate tuple under a coordinate system.

\section{Working with matrices}

\subsection{Multiplying matrices}

Matrix addition is simple:
add the elements at the same positions:
\Formula{
    (AB)_{ij} = A_{ij} + B_{ij}
}

The cell \((AB)_{ij}\) depends on row \(i\) of \(A\) and column \(j\) of \(B\).
\Formula{
    (AB)_{ij} = \sum_k A_{ik} B_{kj}
}

\subsection{Abbreviating a system of linear equations into a matrix equation}

A matrix \enquote{packs} a system of linear equations into one equation.

\[ AB = C \]

\subsection{Linear function? Using matrix to describe a linear transformation?}

A matrix is a linear map?

A matrix is a vector of covectors?

A linear function is called linear because it describes a line.

A function \(f\) is \emph{linear} iff \( f(x+y) = f(x) + f(y) \).

A \emph{transformation} is another word for function.

Transformation.
Linear transformation.
$T(ax+by) = aTx + bTy$.

Matrix arises naturally for describing linear transformations on a vector space.

\section{Generalizing vectors, matrices, and tensors to functions of indexes}

The notation \(A(i)\) suggests that a vector is a function \( I \to \Real \).
The notation \(A(i,j)\) suggests that a matrix is a function \( (I,J) \to \Real \).

\section{Thinking of a matrix as a linear operator}

\section{Tuple vs array vs matrix vs tensor}

An array is a bunch of numbers.

A matrix is a \emph{rectangular} array for which matrix operations are defined.

A tensor carries with it a coordinate system?

A tensor is \emph{not only} a multi-dimensional array of numbers.

A tuple is a one-dimensional (linear) array.

A matrix is a two-dimensional (rectangular) array.

A tensor is a multi-dimensional array.
