\chapter{Nuclear power}

\Hyperlink{https://en.wikipedia.org/wiki/Nuclear_physics}{Nuclear physics, history, modern}

\cite{harms2000principles}

\paragraph{Why study nuclear power?}
We want to build a nuclear fusion power plant.

% Summary
% Question prioritization

% has some references
% http://www.nuclear-power.net/laws-of-conservation/law-of-conservation-of-momentum/
% from google search: nuclear product momentum

% http://www.ppe.gla.ac.uk/~protopop/teaching/NPP/P4-NPP.pdf
% http://www.ppe.gla.ac.uk/~protopop/teaching/NPP/

% https://en.wikipedia.org/wiki/Fusion_power
% https://en.wikipedia.org/wiki/Particle_accelerator
% https://en.wikipedia.org/wiki/Electrostatic_nuclear_accelerator

% https://en.wikipedia.org/wiki/Ab_initio_methods_(nuclear_physics)

% Chemistry 418/518 Nuclear Chemistry Winter 2015
% http://oregonstate.edu/instruct/ch374/ch418518/
% http://oregonstate.edu/instruct/ch374/ch418518/Chapter%2010%20NUCLEAR%20REACTIONS.pdf

% http://www.academia.edu/23818742/Blue_Sky_Research_Fusion_Power_via_Piezoelectric_Crystals

\ResearchQuestion{How practical is maintaining a 5-megavolt electric field gradient to slow down a typical alpha particle?}

\ResearchQuestion{Vacuum breakdown voltage? Why is electrical arc in vacuum even possible?}

What determines the momentum of the fusion products?

Ideas that have been tried?

Who coined the term \emph{radioactive}?

What is a nuclide?

What is an isotope?

What is a nuclear decay?

What is half life?

What is plasma physics?

Heavy atoms split.
Why?

\section{Summary}

We want to have \emph{practically unlimited} energy.

Nuclear power source is the densest known power source in 2017.
% https://en.wikipedia.org/wiki/Space-based_solar_power

A \emph{nucleon} is a proton or a neutron.

\section{Understanding current approaches of building fusion plants}

% https://en.wikipedia.org/wiki/Nuclear_fusion#Methods_for_achieving_fusion

Current projects:
ITER fusion reactor.
% https://www.iter.org/

% https://en.wikipedia.org/wiki/Lawson_criterion
% https://en.wikipedia.org/wiki/Fusion_energy_gain_factor

Crowdfunded nuclear reactor?

\section{Dreaming of personal nuclear power plants}

%https://whatisnuclear.com/
%https://en.wikipedia.org/wiki/Table_of_nuclides
%https://www-nds.iaea.org/relnsd/vcharthtml/VChartHTML.html

Let's build a personal nuclear power plant at home.
We dream of a tool for extracting energy out of the air we breathe.
We dream of something like Mr. Fusion from \enquote{Back to the future} (1985 film).
Put anything in, as long as it is lighter than iron, and the mass transforms into energy.
This may be too na\"ive.

% https://en.wikipedia.org/wiki/Proton%E2%80%93proton_chain_reaction

% https://en.wikipedia.org/wiki/Small,_sealed,_transportable,_autonomous_reactor

Goal:
portable safe personal nuclear reactor.

\section{Understanding the atomic theory and the nucleus}

An \emph{atom} has a \emph{nucleus}.%
\footnote{The word \emph{atom} was borrowed from a Greek word meaning \emph{indivisible}.
The word \emph{nucleus} was borrowed from a Latin word meaning \emph{kernel}.}

A \emph{nucleus} has \emph{protons} and \emph{neutrons}.

The \emph{proton number} is the number of protons.
The \emph{neutron number} is the number of neutrons.
The \emph{mass number} is the sum of the proton number and the neutron number.

An \emph{isotope} has the same proton number but different neutron number.

% https://en.wikipedia.org/wiki/Uranium
A uranium atom (U) has 92 protons.
The isotope U-238 has 146 neutrons because \(92 + 143 = 235\).
\enquote{U-238} means the uranium isotope with mass number 238.

\ExercisePage{
\section{Exercise: Atomic theory}
\ExerciseAnswer{What is the mass number of a U-235 atom?}{235. This is by definition: the mass number of A-xyz is xyz.}
\ExerciseAnswer{How many protons are in a U-235 atom?}{92. This is by definition, so the only way to know it is to remember it.}
\ExerciseAnswer{How many neutrons are in a U-235 atom?}{\(235 - 92 = 143\).
If you remember the proton number of U, you can compute the neutron number of U-235, U-238, and so on.}
\ExerciseAnswer{Is an atom with 100 protons and 120 neutrons an isotope of uranium?}{No. Every uranium atom has 92 protons by definition.}
}

Democritus...

Dalton's theory of atoms...

Farnsworth's fusor is an electrostatic confinement fusion device, but its net power is negative.

\section{Getting used to electronvolt, a unit of energy}

\index{electronvolt}%
\emph{Electronvolt} is a unit of energy:
\begin{equation}
    \SI{1}{eV} \approx \SI{1.6E-19}{J}
\end{equation}

One \emph{electronvolt} is the kinetic energy that an \emph{electron}
gains by climbing one \emph{volt} of electric potential difference.
One electronvolt is the absolute charge of an electron multiplied by one volt.

One electronvolt is the difference of the kinetic energy of an electron
due to moving across one volt of potential difference.

% https://en.wikipedia.org/wiki/Orders_of_magnitude_(energy)
% What? The food energy of a snicker bar exceeds the kinetic energy of a 2-ton vehicle traveling at 115 km/h?
Let's build our intuition of electronvolts.
These are approximate.

\SI{2}{eV} is the energy of a red photon (600 nm wavelength)?

\SI{10}{eV} is ultraviolet-A?

\SI{5}{MeV} is the typical energy of an alpha particle?

\section{Understanding radioactivity}

% https://en.wikipedia.org/wiki/Henri_Becquerel
% https://en.wikipedia.org/wiki/Abel_Ni%C3%A9pce_de_Saint-Victor#Near-discovery_of_radioactivity

What is a photographic plate?

% FIXME copied from Wikipedia
1857:
\enquote{Ni\'epce de Saint-Victor observed that,
even in complete darkness,
certain salts could expose photographic emulsions.}

1896:
Becquerel found that uranium darkens photographic plates.

% Rutherford

% https://en.wikipedia.org/wiki/Radio#Etymology
% animation of an antenna transmitting a radio wave
% https://en.wikipedia.org/wiki/Radio_wave

How do we measure radioactivity?

\section{Nuclear fusion plan?}

Quantum tunneling and fusion?
%https://en.wikipedia.org/wiki/Gamow_factor

Quantum Tunneling in Nuclear Fusion, 1997
%https://arxiv.org/pdf/nucl-th/9708036.pdf

How do we craft the wave function to facilitate fusion?

The sun is massive enough for fusion to just happen.
What's the smallest fusion power plant we can make?

\section{Understanding plasma confinement challenges}

Does fusion power plant require confinement?
Can we build one without confinement?
Can we build one without vacuum possible?

\section{Understanding piezoelectricity}

How does piezoelectricity work?
How does electret work?

Can piezoelectric achieve at least \SI{50}{kV}?
% https://en.wikipedia.org/wiki/Pyroelectric_fusion
% https://en.wikipedia.org/wiki/Cockcroft%E2%80%93Walton_generator
% https://en.wikipedia.org/wiki/Inertial_electrostatic_confinement

What is the electric field strength required to stop a \SI{5}{MeV} alpha particle in distance \(d\)?
Assume that the mass is \SI{6.6E-27}{kg} and the charge is \(2e\).
% https://en.wikipedia.org/wiki/Alpha_particle

\section{Understanding aneutronic fusion}

% https://en.wikipedia.org/wiki/Aneutronic_fusion
Aneutronic fusion is fusion that doesn't produce neutrons.
Neutron is lost energy.

Neutron has zero charge but non-zero magnetic moment.
We can steer a neutron with a magnetic field.

%is this legitimate?
%https://www.indiegogo.com/projects/focus-fusion-empowertheworld--3#/

% https://en.wikipedia.org/wiki/Stellar_nucleosynthesis

Where does the produced energy go?
Some escapes.
Some helps other atoms fuse.

\section{Understanding tokamaks and why they are popular}

What is a tokamak?
Is it popular?
Why?

The beam loses energy.
It radiates them.
Because it is accelerated to move in circles.

Impulsing a pendulum
Back and forth laser oscillation, add power

\section{Understanding how quantum tunneling affects nuclear fusion}

% HOW LOW - ENERGY FUSION CAN OCCUR
% https://arxiv.org/pdf/1211.1243.pdf

Probability of nuclear fusion due to quantum tunneling?

% https://en.wikipedia.org/wiki/Nonlinear_optics#Frequency_doubling
% https://en.wikipedia.org/wiki/Two-photon_absorption
% https://en.wikipedia.org/wiki/Two-photon_physics
There may be a device that absorbs 2 photons with energy \(hf\) each and emits 1 photon with energy \(2hf\).
Frequency doubling.
Optical capacitor?
Optical amplifier?

\section{Research question: Can we focus matter-wave as we focus light waves?}

Quantum parabolic reflector/focuser?
And use that to increase the probability of fusion?

How can we use our knowledge of particle physics to solve the world's energy problem?
Almost everything we can see can be harnessed.
Even the air.

Hybrid fission-fusion reactor, like hybrid fission-fusion bomb?

Quantum electrodynamics, quantum chromodynamics, quantum field theory

% quantum physics
% https://en.m.wikipedia.org/wiki/Matter_wave

\section{Research question: How do we build nuclear fusion reactors?}

stable open-air plasma ring:
piezoelectric plasma generator?
\url{https://futurism.com/stable-plasma-ring-created-open-air-first-time-ever/}

% use electret for plasma confinement for nuclear fusion?
% https://en.wikipedia.org/wiki/Electret

\section{Research question: How do we build self-sustaining plasma?}

% https://www.quora.com/What-are-the-cheapest-metals
% Materials Selection in Mechanical Design, Fifth Edition (9780081005996): Michael F. Ashby: Books

% kurikulum teknik nuklir UGM
% https://tf.ugm.ac.id/images/akademik/Kurikulum_TN_2016.pdf

%Pengantar Teknik Nuklir
%Deteksi dan Pengukuran Radiasi 
%Elektronika Nuklir
%Praktikum Elektronika Nuklir
%Fisika Reaktor Nuklir
%Praktikum Deteksi dan Pengukuran Radiasi
%Radiokimia
%Praktikum Radiokimia
%Proteksi Radiasi
%Komputasi Nuklir
%Prakt ikum Fisika Reaktor Nuklir
%Pengelolaan dan Pengolahan Limbah Radioaktif
%Sistem Keselamatan, Keamanan dan Safeguard Nuklir 

%TKN 4501 Termal Hidraulika Reaktor Nuklir (*) 3 2
%TKN 4502 Teknologi Pembangkit Daya Nuklir (*) 2 3
%TKN 4503 Instrumentasi Nuklir (*#$) 2 4
%TKN 4504 Material Nuklir (#$) 2 5
%TKN 4505 Analisis Reaktor Nuklir (*$)
%Pengelolaan dan Pengolahan BBN (*#$)
%Kimia Radiasi
%Perancangan Sistem Nuklir (*#$)
%Teknologi Reaktor Maju (*#$)
%Sistem Komponen Pendukung Reaktor Nuklir
%Manajemen BBN dalam Teras Reaktor (*#$)
%Sistem Kogenerasi Nuklir
%Teknologi Reaktor Fusi Nuklir (*#$)
%Teknologi Akselerator
%Teknologi Pengendalian Reaktor Nuklir
%Penerapan Radioisotop (*#)
%Penerapan Radiasi (*#)
%Teknik Pemisahan Isotop (*#)

%[1] R.L.  Murray, 2014 .  Nuclear Energy: A n Introduction to The Concepts, Systems, and Application of Nuclear Processes, 7 th Ed .  Butterworth - Heinnem an

Fermi's golden rule:
probability of quantum state transition
% https://en.wikipedia.org/wiki/Fermi's_golden_rule

% https://en.wikipedia.org/wiki/Fusion_power
% A reaction's cross section, denoted σ, is the measure of the probability that a fusion reaction will happen.
% In a plasma, particle velocity can be characterized using a probability distribution.

% https://en.wikipedia.org/wiki/Townsend_discharge
% https://en.wikipedia.org/wiki/Paschen%27s_law
% https://en.wikipedia.org/wiki/Two-photon_physics

\section{Converting other forms of energy to electrical energy}

% https://en.wikipedia.org/wiki/Direct_energy_conversion

Photon-matter interaction

% The interaction of photons with the matter.
% http://meroli.web.cern.ch/meroli/Lecture_photon_interaction.html
% https://physics.stackexchange.com/questions/83105/explain-reflection-laws-at-the-atomic-level
% https://physics.stackexchange.com/questions/132515/does-a-reflection-still-transfer-momentum-to-an-mirror

Photon-photon interaction

\section{Understanding weak and strong interaction}

Flavor change happens by weak decay.

Weak force, beta decay

% https://en.wikipedia.org/wiki/Weak_interaction

Other names: Weak force, weak interaction, weak nuclear force.

Field names:
Quantum flavordynamics / electroweak theory studies weak interaction.
Quantum chromodynamics studies strong interaction.

Strong force binds nucleus together / keeps nucleus intact.
Strong force confines quark into hadron.

\section{Navigating the particle zoo}

The Standard Model continues atomic theory.

% https://en.m.wikipedia.org/wiki/Nuclear_force

A \emph{meson} is 1 quark and 1 antiquark.

A \emph{baryon} is 3 quarks.

Gluon

Baryon

A \emph{fermion} is a particle that follows Fermi\textendash{}Dirac statistics.
A \emph{boson} is a particle that follows Bose\textendash{}Einstein statistics.

Quark

Lepton

Pion

Kaon

Muon

\section{Mass defect and nuclear binding energy: Where does the mass go?}

Revisit Lavoisier's law of conservation of mass.

\( E = mc^2 \).

An extremely tiny mass is \emph{lost} (converted into energy that escapes the weight scale).

Nuclear binding energy

\section{Nuclear fission}

\section{Nuclear fusion}

\section{Finding current research}

IAEA (International Atomic Energy Agency) holds a fusion energy conference every two years.
% https://www.iaea.org/
% http://www-pub.iaea.org/iaeameetings/48315/26th-IAEA-Fusion-Energy-Conference
% https://nucleus.iaea.org/sites/fusionportal/Pages/FEC/FEC.aspx


% Photonuclear reactions triggered by lightning discharge
% https://www.nature.com/articles/nature24630

% Converting chemical energy to electrical energy
% https://en.wikipedia.org/wiki/Fuel_cell

% https://en.wikipedia.org/wiki/Physics_beyond_the_Standard_Model

Criteria and candidates for terrestrial reactions
\footnote{\url{https://en.wikipedia.org/wiki/Nuclear_fusion\#Criteria_and_candidates_for_terrestrial_reactions}}

A common method for detecting neutrons involves converting the energy released from neutron capture reactions into electrical signals.%
\footnote{\url{https://en.wikipedia.org/wiki/Neutron\#Detection}}

\section{Energy}

Energy is a quantity that happens to be conserved?

\section{Wrongish}

Why is there no electron-electron fusion?
Because the strong nuclear force doesn't work on electrons.
Electrons are leptons?
Proton and neutron are baryons?
Hadrons?

\footnote{\url{https://en.wikipedia.org/wiki/Neutron_transport}}

\section{Unclear websites}

\footnote{\url{http://reactor-core.org/}}
