% shared:
% mathematics
% physics

\chapter{Writing}

\epigraph{It seems that perfection is attained not when there is nothing more to add, but when there is nothing more to remove.}{Antoine de Saint-Exup\'ery (1900\textendash{}1944), \emph{L'Avion}}

Writing is a scalable way to spread your ideas.

\section{Reasons for reading fiction}

In fiction, you can pretend that you are someone else.

Fiction helps you imagine.

Fiction helps you escape reality.

\section{Deletion}

Every word must be important.
Use the fewest words to influence the reader.

\section{Respect the reader}

The highest honor for a writer is the reader's attention.

\section{Internet}

Once it's on the Internet, it's there forever.
You can't delete it.

Don't write things that can be used against you.

\section{How to write a book}

Write things.
Order things.
Delete as many as possible.
Repeat.

Alternate between focusedly-dip-deep-into-a-section and be-a-generalist-and-read-a-chapter.

A book begins as a bunch of small isolated islands of text containing main ideas and structural drafts.

Write.
Make mistakes.
Rewrite.

\section{Form}

The first sentence of a paragraph makes claim.
The rest of the paragraph supports the claim.
Thus, the reader can skim by reading the first sentence of every paragraph.

\emph{Ordering}:
If understanding \(B\) depends on understanding \(A\),
then \(A\) must come before \(B\) in the text.
Don't make the reader jump around.

Define uncommon term before using it.

\emph{Grouping}:
Related text should be near.

\emph{Non-redundancy}:
Never waste what the reader has read.

\emph{Clarity}:
If the reader reads the book linearly,
then he/she should understand a sentence just by reading it once.
The reader should parse the sentence in one pass.
Avoid ambiguous syntax.

% https://en.wikipedia.org/wiki/Garden_path_sentence
Avoid \emph{garden path sentences}.
Readers must parse a sentence before they understand it.
Understanding is hard.
Parsing adds unnecessary difficulty.

\section{Why a book, not a website?}

A book has pages.
Pages help readers estimate the amount of material.
Pages help readers estimate the amount of time they have to invest.
Pages are significant visual cue.
Online text seems endless.
Links distract.

\section{Academic publishing reform}

Do not let Aaron Swartz die in vain.

% https://en.wikipedia.org/wiki/Aaron_Swartz#Open_Access
% https://en.wikisource.org/wiki/Guerilla_Open_Access_Manifesto
% https://archive.org/stream/GuerillaOpenAccessManifesto/Goamjuly2008_djvu.txt

% https://en.wikipedia.org/wiki/Library_Genesis

% https://en.wikipedia.org/wiki/ICanHazPDF

% https://en.wikipedia.org/wiki/Sci-Hub
% https://vk.com/sci_hub

% https://en.wikipedia.org/wiki/Academic_journal_publishing_reform

% https://en.wikipedia.org/wiki/Library.nu

\section{Open-access journals}

% http://www.mdpi.com/journal/universe

% https://benthamopen.com/PHY/home/

After reading \cite{gopen1990science}, we conclude:
The subjects of the sentences in a paragraph should match the paragraph's topic.
IF passivating a sentence would match its subject to the topic of the containing paragraph, then passivate it.
Don't mindlessly activate all sentences.

There is a summary%
\footnote{\url{https://www.crowl.org/Lawrence/writing/GopenSwan90.html}}.
