\chapter{Conjugables}

\section{I-adjective}

黒いblack

細いthin

\section{Existence and possession: 有る、在る、居る}

有る（ある）possess

在る（ある）exist (inanimate)

居る（いる）exist (animate)

リンゴが有る。
The implied entity has an apple.

リンゴが在る。
``There is an apple,''
or ``An apple exists,''
depending on context.

犬は居る。
There is a dog.

妹が居る。
The implied entity has a younger sister.

\section{Negation: 無い}

Irregular:

ある becomes 無い（ない）: not; not exist

Regular:

v1-ru-to-nai

居る becomes 居ない.

食べる becomes 食べない.

v5-u-to-anai

笑う becomes 笑わない.

Examples:

リンゴが無い。
``The implied entity does not have any apples,''
or ``There are no apples,''
depending on context.

妹が居ない。
The implied entity does not have a younger sister.

\section{Perfection of dictionary-form verb: た、だ}

``Perfect'' means ``complete'', as in ``past perfect'', not flawless.
Every dictionary-form verb can be perfected into an perfect-form verb.
The result is still a verb.

笑う。
The implied entity laughs.
The implied entity will laugh.

笑った。
The implied entity laughed (and is no longer laughing now).
The implied entity has laughed.
The implied entity had laughed.

\section{Perfection of i-adjective: i-to-katta}

速い becomes 速かった

\section{Negation of i-adjective: i-to-kunai}

速い becomes 速くない

\section{i-to-kunakatta = i-to-kunai + i-to-katta}

\section{Adverbiation: i-to-ku}

速い becomes 速く

楽しく歩いたwalked happily

\section{Volitionalization: u-to-itai}

笑う becomes 笑いたい

話す becomes 話したい

\section{Conjunction: i-form}

食べる becomes 食べ

話す becomes 話し

立つ becomes 立ち

\section{Concurrency: X-while-Y: i-form of Y + nagara + X}

Japanese is consistently head-final,
and while-Y explains X,
so while-Y comes before X.
The modifier always comes before the head.

食べながら while eating

立ちながら while standing

食べながら話すto speak while eating

立ちながら食べるto eat while standing

\section{In doubt}

A verbal can be modified by adverbials.
A nominal can be modified by adjectivals.
