\documentclass[12pt,openany]{book}
\usepackage{tabularx}
\usepackage[CJKmath=true]{xeCJK}
\xeCJKDeclareSubCJKBlock{RadicalB}{"20000 -> "2A6DF}
\setCJKmainfont{HanaMinA}
\setCJKmainfont[RadicalB]{HanaMinB}
\usepackage{ruby}
\newcommand\foreign[1]{\textsl{#1}}
\renewcommand\rubysize{0.5}
\renewcommand\rubysep{0.0em}
\usepackage{amsmath}
\usepackage[
    paperwidth=6in,paperheight=9in
    ,top=0.75in,bottom=0.75in
    ,inner=0.67in,outer=0.33in
]{geometry}
\usepackage{hyperref}
\begin{document}
\frontmatter
\phantomsection
\addcontentsline{toc}{chapter}{Cover}
\begin{titlepage}
{%
    \setlength\parindent{0em}%
    {\Large Cold-Turkey Head-First Dive into Kanji}\par\vspace{2em}
    Erik Dominikus\par
    2017-04-02\par
    (Draft)\par\vfill
    {\small
        This book aims to optimize your kanji learning loop.

        This book does not discuss about grammar, register, and culture.
    }
}
\end{titlepage}
\phantomsection
\addcontentsline{toc}{chapter}{\contentsname}
\tableofcontents
\mainmatter
\chapter{Strategy}

\section{Mindful consistent spaced repetition}

The key to remembering kanji is
\emph{mindful consistent spaced repetition}.
Don't just gloss over the kanji.
\emph{Internalize}.

Don't read one stroke at a time,
but read one component at a time.

\emph{Expect to forget a lot, many times}.
You will forget.
Reread this book.
This book is meant to be read and read again many times.
The aim of this book is to help you optimize your loop.

Mindful consistent spaced repetition is \emph{the only way},
at least until mind-upload is invented.
There is no shortcut.
It's just how the brain works.

The loop:
For each kanji:
Read the kanji out (voice it out).

\section{Target audience}

You read, write, speak, and listen to English but don't read, write, speak, and listen to Japanese.

\section{Memorize the syllabaries}

Memorize all hiragana and katakana.
It's a must.

\section{Understand the principles}

Most kanjis are composed using other kanjis.

\section{Read native writers and hear native speakers}

Go to YouTube.
Hear how they speak.

Use Google Translate, but do not trust it too much.
Sometimes it gives the wrong reading.
Corroborate with Wiktionary and Wikipedia.

Find Japanese people on Twitter and understand their tweets.

Find Japanese blog posts.

Set up an input method, or use Google Translate.

\newcommand\myspacing{0.5em}
% side-by-side
\newcommand\sbs[2]{%
    {%
        \parindent0em%
        \par%
        \vspace{\myspacing}%
        \begin{tabularx}{\linewidth}{XX}%
            {\fontsize{12pt}{12pt}\selectfont#1}%
            &%
            {\fontsize{10pt}{12pt}\selectfont#2}%
        \end{tabularx}%
        \vspace{\myspacing}%
        \par%
    }%
}

\input{kanji}
\input{kanji-1}
\input{kanji-2}
\input{kanji-3}
\input{kanji-4}
\input{kanji-5}
\input{kanji-6}
\input{kanji-7}
\input{kanji-8}
\input{kanji-9}
\input{kanji-10}
\input{kanji-11}
\input{kanji-12}
\input{kanji-13}
\input{trap}
\input{grammar-1}
\input{conjugable}
\input{nominal}
\input{statement}
\input{grammar-2}
\input{grammar-3}
\input{polite}
\input{logic}
\input{scratch}
\input{grammar}

\end{document}
