\section{People-related}

\subsection{2 人 3 己大士女 4 犬太父 5 立母兄民 6 共}

人（ジン、レン、ニン、ひと）person; people; human

己（コ、おのれ）self

大 depicts a person with outstretched arms.
大（タイ）big.
大きい（おおきい）big.

士（さむらい）gentleman; samurai.
士（シ）(suffix)
a qualified person;
a person with a qualified profession.
学士（ガクシ）university graduate (who has obtained a degree).
工学士（コウガクシ）Bachelor of Engineering.

女（おんな）woman

犬（いぬ）dog.

太い（タイ、タ、ふとい）(adj-i) fat; thick.
太る（ふとる）to grow fat; to become fat; to gain weight.

父（フ、ちち、とう）father.
父（ちち）(humble) father.
お父さん（おとうさん）(honorific) father.

立 depicts a person standing on ground.
立つ（たつ）(vi) to stand

母 depicts a pair of breasts.
母（ボ、はは、かあ）mother.
母（はは）(humble) mother.
お母さん（おかあさん）(honorific) mother.

兄（ケイ、キョウ、あに）older brother.
兄（あに）(humble) older brother.
お兄さん（おにいさん）(honorific) older brother.

民（ミン、たみ）people

共（とも）companion; follower; attendant; retinue.

\subsection{4 天夫 5 矢失}

天（テン）sky; heaven

夫（フ、おっと）husband

矢（シ、や）arrow

失 depicts something falling from a hand.
失う（うしなう）(vt) to lose; to part with.
失明（シツメイ）loss of eyesight.
失血（シッケツ）loss of blood.

\subsection{2 了 3 子 6 字 7 呆弟男児}

了 depicts a wrapped baby with only head visible.

子 depicts a child with arms visible.
子（シ、こ） child.
男子（ダンシ）young man, boy.
女子（ジョシ）young woman, girl.
太子（タイシ）crown prince.

字（ジ）character, letter (of an alphabet, not the letter that is a document)
赤字（あかジ）deficit; red letter; red Han-character.
漢字（カンジ）Han character.

呆 depicts a child.
呆れる（あきれる）(v1,vi) to be amazed, astonished, astounded.

弟（テイ、ダイ、おとうと）younger brother.
弟（おとうと）(humble) younger brother.
弟さん（おとうとさん）(honorific) younger brother.
兄弟（キョウダイ）siblings;
brothers and sisters
(although the characters mean older brother and younger brother).

男（おとこ）man.

児 depicts an infant with imperfect cranium (fontanelles).
児 is simplified from 兒.
乳児（ニュウジ）infant; suckling baby.
男児（ダンジ）boy; son.

\subsection{Body parts: 2 又 3 口 4 心氏止手毛开 5 目 6 自血耳舌 9 首}

又 depicts the right hand.
又（また）again; also.

口（くち）mouth.

心（シン、こころ）heart.

氏 depicts a man bowing to the left.
氏（シ）(suffix,honorific) Mr.; Mrs.. family. clan.
氏（うじ）family name; birth; lineage.

止 depicts a footprint.
止（シ、と）stop.
止まる（とまる）(vi) to stop (moving); to come to a halt.
止める（やめる）(v1) to stop (doing something).
止める（とめる）(v1) to stop moving (walking, etc.); to park (a car).

手（て）hand.
手がける（てがける）(v1) to make; to do; to produce; to work on.
手洗い（てあらい）restroom; lavatory; toilet; a place for washing hands

毛（モウ、け）hair.

开 is a simplification of 幵 (raising both hands).

目（め）eye

自（ジ、シ、みずか）self; oneself.
自ら（みずから）(adv) personally.
自在（ジザイ）freely (at will).
自分（ジブン）self.
自身（ジシン）self.

血（ち）blood.
止血（シケツ）stop bleeding; hemostasis.

耳（みみ）ear

舌（ゼツ、した）tongue.

首（シュ、くび）neck

\subsection{Blades: 2 刀刂 3 弋 4 戈斤 5 戊 6 戌 9 咸}

刀（トン、かたな）sword.

刂 depicts a knife.

弋 depicts shooting with a bow and an arrow.

戈 depicts a spear-axe (a halberd).

斤 depicts an axe.

戊 depicts a dagger-axe (an ancient Chinese weapon).

戌 depicts an axe.

咸 depicts an axe and a mouth.

\subsection{Administrative divisions: 4 区 5 市 7 町 8 京 9 県}

% https://en.wikipedia.org/wiki/Administrative_divisions_of_Japan

区（ク）ward; district (an administrative division).

市（シ）city (an administrative division).
市（いち）market; fair (trade show).

町（チョウ、まち）town.

京（キョウ、ケイ、みやこ）.
京（キョウ）imperial capital.

県（ケン）prefecture.

\subsection{4 予 7 序}

予（ヨ、あらかじ）beforehand.
予め（あらかじめ）beforehand.

序（ジョ）preface

\subsection{4 介}

介（カイ）can mean mediation or shellfish.
紹介（ショウカイ）introduction; referral.
仲介（チュウカイ）agency; intermediation.
介入（カイニュウ）intervention.
介助（カイジョ）help; assistance; aid.
魚介（ギョカイ）marine products; seafood.

\subsection{4 今 6 合会}

今（コン、いま）now.
今回（コンカイ）this time; this occasion; this occurrence.

合（ゴウ、あ）fit.
合う（あう）to fit.

会（カイ、エ、あ）meet.
会う（あう）to meet (face to face).
会社（カイシャ）company; corporation.
会社員（カイシャイン）company employee.
年会（ネンカイ）yearly meeting; annual convention.
社会（シャカイ）society.
会見（カイケン）interview.

\subsection{2 丁 5 庁}

丁（ひのと）depicts a nail.
丁（チョウ）(suffix)
counter for long and narrow (relatively flat) thing
such as paper, guns, scissors, spades, hoes, guitars.
一丁（イッチョウ）one sheet; one page.
one serving (in a restaurant).
丁重（テイチョウ）polite; courteous; hospitable.
包丁（ホウチョウ）kitchen knife.

丁（チョウ）town section.

庁（チョウ）government office.

\subsection{2 力}

力（リョク、リキ、ちから）strength.
百人力（ヒャクジンリキ）tremendous strength.

\subsection{7 貝 9 頁}

貝 depicts a cowry (a kind of seashell used as money in ancient China).
貝（かい）shell; shellfish.

頁 originally depicts a head,
but comes to mean ``leaf'' (an organ in plants).

\subsection{7 図車声}

図（ズ）drawing; map; picture; plan; illustration; diagram; figure; chart.
図る（はかる）to plot; to plan; to design.

車（くるま）car; vehicle. wheel.
水車（スイシャ）water wheel.

声（セイ、こえ）voice.
This character was simplified from 17 聲.

\subsection{8 門 9 重面 10 書 12 傘}

門（モン、かど） gate

重 depicts a man carrying a bag.
重い（おもい）heavy (of weight).

面（おもて）mask.
面白い（おもしろい）interesting.

書 depicts a hand holding a pen writing on paper.
書（ショ）writing.
辞書（ジショ）dictionary.
書く（かく）to write.

傘（かさ）umbrella.
