\chapter{Kanji 5}

\section{5 市}

市（シ）city (an administrative division).

市（いち）market; fair (trade show).

\section{5 矛务}

矛 Kangxi radical 110 (``spear'').

\subsection{11 務}

務（ム）duty.

事務所（ジムショ）office.

\section{5 乎平半 8 呼}

乎 question mark.

平（ヘイ）flat.

平たい（ひらたい）(adj-i) flat; even; level; simple.

平ら（たいら）flatness.

平皿（ひらざら）flat dish.

水平（スイヘイ）level; horizontally.

平面（ヘイメン）level (flat and not-tilted) surface.

平安（ヘイアン）peace; tranquility.

平気（ヘイキ）coolness; calmness; composure; unconcern.

公平（コウヘイ）fairness; impartiality; justice.

平日（ヘイジツ）weekday; ordinary day (non-holiday).

平年（ヘイネン）normal (non-leap) year; normal year (related to harvest; weather).

平文（ヘイブン）plain (non-encrypted) text.

半（ハン）half.

半ば（なかば）half; middle; semi.

呼（コ）call.

呼ぶ（よぶ）to call; to invoke; to summon.

\section{5 乍 7 作 9 昨}

作（サク）make; work; harvest.

作る（つくる）to make.

昨（サク）previous.

昨日（きのう）yesterday.

\section{5 㠯 8 官 9 追}

官 depicts many rooms in a building.

官（カン）government official.

官界（カンカイ）bureaucracy.

長官（チョウカン）secretary; director; chief; director general.

追（ツイ、お）chase; follow; pursue.

追う（おう）to chase; to run after; to follow.

追い風（おいかぜ）tailwind; favorable wind.

\subsection{7 呂 12 営}

呂（ロ）spine.

営（エイ）manage.

\section{5 可司 6 向同回 10 高}

可（カ）passable; acceptable; tolerable.

許可（キョカ）permission; authorization; approval.

司（シ）director.

司る（つかさどる）to rule; to govern.

向 depicts a house and a window.

向かい（むかい）(n) facing; opposite; across the street; other side.

向く（むく）to face; to turn toward.

向ける（むける）(v1,vt) to turn towards.

向こう（むこう）opposite side; other side; opposite direction.

向上（コウジョウ）improvement; advancement; progress.

同（ドウ、おな）same.

同じ（おなじ）same.

同性愛（ドウセイアイ）same-sex love.

回 depicts a spiral.

回す（まわす）(vt) to turn; to rotate.

今回（コンカイ）this time.

一回（イッカイ）once; one time.

一回目（イッカイめ）first.

高（コウ）high.

高い（たかい）high; expensive.

\subsection{7 何 8 河}

何（カ、なに、なん）what.

河（かわ）river; stream.

河川（カセン）rivers.

大河（タイガ）large river.

\subsection{11 寄}

寄（キ）approach.

寄り（より）(n) being close to.

寄せる（よせる）(v1,vt) to come near.

寄り掛かる（よりかかる）to lean against; to lean on.

\ruby{柵}{さく}に寄りかかるto lean on the fence.

\subsection{10 哥 14 歌}

哥 means older brother.

歌（カ、うた）song.

歌声（うたごえ）singing voice.

歌う（うたう）to sing.

\subsection{13 稿}

稿（コウ）draft; copy; manuscript

\section{5 业}

业 is an alternative form of 北 (north).

\subsection{8 並 12 普}

並 depicts two people standing side-by-side.

並（ヘイ、なみ、なら）row.

並（なみ）row.

並ぶ（ならぶ）to stand in line; to line up.

普（フ）universal; wide; general.

普通（フツウ）(adj-no) general; ordinary; usual.

普通の人間（フツウのニンゲン）ordinary human.

\subsection{13 業 14 菐僕}

業（ギョウ）industry.

工業（コウギョウ）manufacturing industry.

企業（キギョウ）enterprise; business.

菐 is the 14-stroke ``thicket'' radical.

僕（ボク）I; me (male).

\section{5 左右 8 若}

左（ひだり）left.

右（みぎ）right.

若い（わかい）young; at an early time in life.
若年（ジャクネン）the time when one was young.

\section{5 玉 8 咅}

玉（ギョク、たま） ball

玉を\ruby{投}{な}げて！Throw the ball!

咅 means ``to spit out''.

\subsection{6 全}

全 depicts a whole piece of jade.

全（ゼン）whole.

全国（ゼンコク）countrywide; national.

全て（すべて）all; everything.

全く（まったく）(adv) completely, entirely, wholly, totally

\subsection{10 倍剖 11 部}

倍（バイ）double.

3倍（サンバイ）three times; threefold.

剖（ボウ）dissection.

部（ブ）section; department; part.

市部（シブ）urban areas.

全部（ゼンブ）all; entire; whole.

学部（ガクブ）a department in a faculty in a university.

東部（トウブ）eastern part.

部門（ブモン）division (of a larger group).

一部（イチブ）one copy (of a document).

\section{5 主 7 住 8 往注}

主（おも）chief; main; principal; important.

主人公（シュジンコウ）hero; main character.

住（ジュウ）dwelling; living.

永住（エイジュウ）permanent residence.

居住（キョジュウ）residence.

住人（ジュウニン）inhabitant; resident; dweller.

住む（すむ）to live (of humans); to reside; to inhabit.

日本に住んでいるto be living/residing in Japan.

往（オウ）outward; journey.

注（チュウ）pour.

注文（チュウモン）order (in a restaurant); request.

\section{5 以}

以（イ）by means of.

以前（イゼン）previously; formerly; the former.

以上（イジョウ）above; more.

１０以上above 10; over 10; more than 10.

以下（イカ）below; less.

１０以下below 10; under 10; less than 10.

\section{5 去 8 法}

去（キョ、コ）leave

法（ホウ）method.

\section{5 令 (6 米) 7 冷}

令（レイ）orders

冷たい（つめたい）(adj-i) cold (of a tangible object; to the touch)

\subsection{8 命}

命（メイ、ミョウ、いのち）fate.

運命（ウンメイ）fate; destiny.

使命（シメイ）mission.

\subsection{12 奥}

奥（おく）interior.

奥山（おくやま）remote mountain.

\subsection{12 歯 17 齢}

歯（シ、は）tooth.

歯車（はぐるま）gear; cog-wheel.

齢（レイ）age.

齢は、２３\ruby{歳}{サイ}。The age is 23 years.

\section{5 号}

号（ゴウ、よびな、さけ）number

\section{5 皿且 6 血}

皿（さら）dish, plate

且つ（かつ）and.

且又（かつまた）besides; furthermore; moreover

血（ち）blood.

止血（シケツ）stop bleeding; hemostasis.

\section{5 示}

示 depicts a spirit.
There are two Unicode codepoints:
⽰ (U+2F70 in the CJK Radicals Supplement block)
and 示 (U+793A in the CJK Unified Ideographs block).
To machines they differ,
but to humans they look the same.

示（シ）indicate.

示す（しめす）to indicate.

\subsection{13 禁}

禁（キン）prohibition.

禁じる（キンじる）(v1,vt) to prohibit.

禁止（キンシ）ban.

\section{5 世台}

世（よ）world; society; age; generation

台（ダイ、タイ、うてな）tower; stand; pedestal.
台 is simplified form of 14 臺.
仙台（センダイ）(city name) Sendai.

\section{5 冬 10 夏 11 終}

冬（トウ、ふゆ）winter

夏（カ、なつ）summer

終（シュウ）end.

最終（サイシュウ）last; final; closing.

終了（シュウリョウ）(n,vs)
end; close; termination.
to quit or exit (a computer program).

終わる（おわる）to finish; to end; to close.

\section{5 犯 8 狙 9 狭}

犯（ハン、おか）crime.
犯す（おかす）to commit (a crime); to perpetrate (a crime).

狙う（ねらう）(vt) to aim at

狭める（せばめる）(v1,vt) to narrow.
狭い（せまい）(adj-i) narrow; confined; small.

\section{5 癶 7 豆}

癶 depicts footsteps, dotted tent, or legs.

豆（トウ、まめ）beans; pea.

\subsection{9 発 12 廃}

発（ハツ、ホツ）departure.

発車（ハッシャ）(n,vs) departure of a vehicle.

発達（ハッタツ）development; growth.

発言（ハツゲン）utterance; speech; proposal.

発信 (ハッシン) (n,vs) dispatch; transmission; submission.

廃（ハイ）abolish.

廃人（ハイジン）crippled/disabled/invalid person.

\subsection{12 登}

登（トウ）climb.

\section{5 史 6 吏 7 更 8 使 9 便}

史（シ）history.

史家（シカ）historian.

吏（リ）officer.

更（コウ）grow late.

更ける（ふける）(vi) to get late; to advance; to wear on.

使（シ）use.

使用（シヨウ）(n) use.

使う（つかう）to use.

便（ベン、ビン）convenience.

便利（ベンリ）convenient; handy; useful.

不便（フベン）inconvenience.

便所（ベンジョ）lavatory.

大便（ダイベン）feces; excrement; shit.

便り（たより）news; tidings; information; letter.

\section{5 召 8 沼招 9 昭 11 紹}

召す（めす）(honorific) to invite; to eat

沼（ぬま）marsh; swamp; bog

招（ショウ）beckon.

招待（ショウタイ）invitation.

昭（ショウ）shining.

紹（ショウ）introduce.

紹介（ショウカイ）introduction; referral.

\section{5 目 6 自}

目（め）eye

自（ジ、シ、みずか）self; oneself.

自ら（みずから）(adv) personally.

自在（ジザイ）freely (at will).

自分（ジブン）self.

自身（ジシン）self.

\subsection{9 相省冒 11 眼}

相（ソウ、あい）mutual.
相手（あいて）companion; partner; company.

相（ショウ）minister.
首相（シュショウ）prime minister; chancellor; premier.

省（ショウ）(suf) ministry; department.
国交省（コッコウショウ）(abbr)
Ministry of Land, Infrastructure, Transport, and Tourism.
省く（はぶく）(vt)
to omit; to leave out; to exclude.
to curtail; to save; to cut down; to economize.

冒 depicts a hat obstructing the sight, implying rashness
(acting without enough thought).
冒す（おかす）(vt) to risk.
冒険（ボウケン）adventure.

眼（まなこ）eyeball.
両眼（リョウガン）both eyes.

\subsection{17 瞳}

瞳（ひとみ）pupil (of the eye).

\subsection{7 見 11 現 12 覚}

見る（みる）(v1,vt) to see.

見える（みえる）(v1,vi) to appear.

現（ゲン）appear.

表現（ヒョウゲン）expression; presentation.

現す（あらわす）(vt) to reveal; to show; to display.

現れる（あらわれる）(v1,vi) to appear; to become visible; to materialize.

覚める（さめる）(v1,vi) to wake up
