\chapter{Other characters}

\section{2 刀（トウ）sword}

刀（かたな）katana (Japanese sword)

\subsection{4 切 (spoon and sword)}

切る（きる）(v5r) cut

\subsection{4 方（かた）(tip of a sword)}

\section{3 女（おんな）woman}

6 好き（すき）like

13 嫌い（きらい）hate

\section{3 山（サン、やま） mountain}

山（やま）mount, mountain

\section{3 工（コウ、ク） craft}

工学（コウガク）engineering

工業（コウギョウ）manufacturing industry

工作（コウサク）work; construction; handicraft

大工（ダイク）carpenter

工（たくみ）(name) Takumi

\section{3 小（ショウ）small}

(three sand granules)

小さい（ちいさい）small

\subsection{4 少 few}

少し（すこし）(adv, n) small quantity; few; a little

止める（やめる）to stop

止める（とめる）to stop

歩く（あるく）to walk

\section{3 土（つち） earth; soil}

9 室（むろ）room (the character consists of roof, private, and earth)

\section{4 中（チュウ、なか） middle}

中学（チュウガク） middle school; junior high school

…の中（…のなか）middle of something

田中（たなか）(name) Tanaka

中山（なかやま）(name) Nakayama

中川（なかがわ）(name) Nakagawa

\section{4 王 king}

\subsection{5 玉（ギョク、たま） ball}

\subsection{8 国（コク、くに） country}

\subsection{9 美 beauty}

美味しい（おいしい）delicious (idiosyncratic reading)

美しい（うつくしい）beautiful

\section{4 天（テン） sky; heaven}

天気（テンキ）weather

天才（テンサイ）genius; prodigy

\section{4 火（カ、ひ）fire}

The fire radical becomes four dot strokes at the bottom side.

4 火（ひ）fire, flame

火事（カジ）fire (disaster).
ラメン店で火事fire at a ramen shop.

\subsection{8 炎（ほのお）flame, blaze}

炎天（エンテン）scorching sun

\subsection{熱い（あつい）(adj)hot (temperature)}

\subsection{燃える（もえる）(v1,vi) to burn}

\section{4 斤（キン） (axe)}

\subsection{13 新（シン） new}

(cut tree down with an axe)

新しい（あたらしい）new

新聞（シンブン）newspaper (lit. new hear)

新車（シンシャ）new car

\section{5 母 (breasts)}

母depicts a pair of breasts.

母（はは）mother

\subsection{7 毎（マイ） every}

毎日（マイニチ）everyday

毎月（マイゲツ、マイつき）every month

毎時（マイジ）every hour

毎回（マイカイ）every time

毎年（マイネン、マイとし）every year

\subsection{8 海（うみ）(water breast)sea}

\section{5 生 (sprout)}

生depicts something sprouting from the ground.

5 生（セイ）nature; sex; gender

生まれる（うまれる）(v1,vi) to be born

8 性（セイ）nature; sex; gender

男性（だんせい）male

女性（じょせい）female

\section{5 目 eye}

目（め）eye

見る（みる）(v1,vt) to see

見える（みえる）(v1,vi) to appear

\section{5 田 rice field}

\subsection{7 男（おとこ）man}

男is田(rice field) and 力(strength).

\subsection{9 界（カイ）world}

学界（ガッカイ）academic world

世界（セカイ）world

業界（ギョウカイ）industry world, business world

\subsection{6 自（ジ） self}

自ら（みずから）(adv) personally

自在（ジザイ）freely (at will)

自分（ジブン）self (context? example usage?)

自爆（ジバク）suicide bombing, self-destruct (lit. self explode)

自動車（ジドウシャ）automobile (lit. self-moving carriage)

\section{6 手（て） hand}

\section{6 回（カイ） (spiral)}

回す（まわす）to turn; to rotate

一回（イッカイ）once; one time

一回目（イッカイめ）first

\section{6 全（ゼン） whole}

(a whole piece of jade)

全部（ゼンブ）altogether; everything

全く（まったく）(adv) completely, entirely, wholly, totally

\section{6 外（ガイ） foreign}

外人（ガイジン）foreigner, foreign person

外国（ガイコク）foreign country

外界（ガイカイ）outside world

次（つぎ）next (in sequence) (?)

\section{6 西（にし） west}

\section{7 形（かたち） shape; form}

\section{8 金（キン、かね）gold; money}

金has a lot to do with metals.

金属（キンゾク）metal. This is also the chemistry term.

重金属（ジュウキンゾク）heavy metal. This is also the chemistry term.

金色（キンいろ）golden (color)

金色（コンジキ）golden (color)

13 鉄（テツ、くろがね）iron (lit. 黒金 black metal)

14 銀（ギン、しろがね）silver (lit. 白金 white metal)

14 銅（ドウ、あかがね）copper (lit. 赤金 red metal)

16 鋼（コウ、はがね）steel

青銅（セイドウ）bronze

鋼鉄（コウテツ）steel

\section{8 門（かど） gate}

\subsection{12 開 open}

開く（ひらく）to open (door, business, eye, mouth, ...)

\subsection{12 間 interval}

\section{9 食（ショク） food, eat}

食べ物（たべもの）food

食物（ショクもの）food

食べる（たべる）(v1) to eat

飲む（のむ）to drink (any liquid, not just liquor)

\section{12 無（ム） no, -less, without}

無駄（ムダ）uselessness

無用（ムヨウ）uselessness

無敵（ムテキ）invincible, unrivaled (lit. no-enemy)

無茶（ムチャ）absurd, unreasonable (lit. no-tea)

無人（ムジン）unmanned (lit. no-human)

無言（ムゴン）silence (lit. no-say)

\section{6 再（サイ） again, re-}

再生（サイセイ）playback; rebirth

再開（サイカイ）reopening

再来（サイライ）return, comeback

\section{6 曲（キョク）music}

作曲（さっきょく）music

音楽（おんがく）music

楽しい（たのしい）happy

了解（りょうかい）understanding; roger that

\section{6 耳（みみ） ear}

14 聞 hear

聞く（きく）to hear

\section{8 隹 (short-tailed bird)}

誰 （だれ）who

\section{10 書（ショ）write}

It's a hand holding a pen writing on paper.

Is this related to 事（ごと）?

書く（かく）to write

\section{12 最（サイ） most}

最も（もっとも）most.
日本の最も高い山（ニホンのもっともたかいやま）Japan's highest mountain.
世界で最も太い人（セカイでもっともふといひと）The fattest person in the world.

最小（サイショウ）smallest

最大（サイダイ）biggest

最初（サイショ）first

最後（サイゴ）last

最新（サイシン）newest

最高（サイコウ）best, highest, tallest

\section{7 貝（かい）cowry}

(Cowry is a kind of seashell used as money in ancient China.)

買う（かう）to buy; to purchase

賣る（うる）to sell.
This kanji has been simplified to 売る.

\section{4 殳 (a hand holding a tool; weapon)}

This appears in
殺す（ころす）(to kill),
電撃（デンゲキ）(electric shock),
投げる（なげる）(to throw),
and 設ける（もうける）(to establish).

\section{6 死（シ）death}

死亡（シボウ）death; mortality

\section{8 使（シ）use}

使う（つかう）to use

使用（シヨウ）(n) use

\section{8 店（テン） (n) store; shop}

ラメン店ramen shop (ramen is a kind of Japanese noodle)

\section{15 暴（バク）(antler of a buck; savage attack)}

暴れる（あばれる）(v1,vi) to rage; to act violently

\subsection{19 爆（バク）burst; explode; bomb}

自爆（ジバク）suicide bombing

水爆（スイバク）hydrogen bomb

原爆（ゲンバク）atomic bomb; nuclear bomb

空爆（クウバク）aerial bombing; air raid

爆殺（バクサツ）killing by bombing

爆死（バクシ）death by explosion

\section{Counting}

5 半（ハン）half

6 百（ヒャク）hundred

3 千（セン）thousand

3 万（マン）ten thousand

\section{Animals}

動物（ドウブツ）animals.
The character says ``moving product''.

牛（ギュウ、うし）
cow; bull; ox; buffalo.
beef.

物（ブツ）.
Semantic 牛(cow) and phonetic 勿（ブツ）.

動物園（ドウブツエン）zoo (lit. animal garden)

4 犬（いぬ）dog.

10 馬（うま）horse.

11 鳥（とり）bird; chicken; chicken meat.
焼き鳥（やきとり）grilled chicken meat.

11 猫（ねこ）cat.

11 豚（ぶた）pig.

\section{Grass}

7 花（はな）flower

9 草（くさ）grass

\section{Personal data}

出身（シュッシン）person's origin (town, city, country, etc.)

出身地（シュッシンチ）birthplace

誕生日（タンジョウビ）birthday; birth date; day of birth.

身長（シンチョウ）height (of body)

体重（タイジュウ）body weight

血液型（ケツエキガタ）blood type.「Ａ型」type a.

好きなもの（すきなもの）likes

嫌いなもの（きらいなもの）dislikes

\section{Hard-to-translate}

さすが

\section{4 反（ハン）anti-}

反する（ハンする）to oppose; to rebel; to revolt

反体制（ハンタイセイ）anti-establishment

\section{8 長（チョウ）long (distance or time); leader}

長い（ながい）long (distance); long (time)

身長（シンチョウ）height (of body)

最長（サイチョウ）longest, tallest

社長（シャチョウ）company president

\section{変（ヘン）strange}

変態（ヘンタイ）sexual perversion

変身（ヘンシン）metamorphosis; transformation

\section{3 幺 (tiny, small, short thread)}

幻（まぼろし）phantom; vision; illusion; dream

\section{6 糸（いと） thread}

糸combines幺and小.

細い（ほそい）thin; slender
The left-side part of 細is the left-side form of 糸.

紅（くれない）deep red; crimson

\section{Axe radical}

axe

\subsection{咸 (mouth and axe)}

減（ゲン）reduction; 10\%減 ten percent reduction.

\section{Country names}

米穀（ベイコク）United States of America

\section{Weekdays and calendars}

These kanji readings for today, yesterday, and tomorrow are irregular.

今日（きょう）today

昨日（きのう）yesterday

明日（あした）tomorrow

Names of weekdays.

日曜日（ニチヨウび）Sunday

月曜日（ゲツヨウび）Monday

火曜日（カヨウび）Tuesday

水曜日（スイヨウび）Wednesday

木曜日（モクヨウび）Thursday

金曜日（キンヨウび）Friday

土曜日（ドヨウび）Saturday

毎日（マイニチ）everyday

Expressions.

また明日（あした）see you again tomorrow; means 'again' and 'tomorrow'

\section{6 羽（はね）feather}

10 弱

弱い（よわい）weak

11 習 (the bottom character is自not日).

習う（ならう）(vt) to learn

練習（レンシュウ）training; practice

\section{Family}

TODO inner-outer

父（ちち）father

母（はは）mother

兄（あに）older brother

姉（あね）older sister
