\chapter{Kanji 3}

\section{3 廾 4 开升}

廾 depicts two hands.

开 is a simplification of 幵 (raising both hands).

升（ショウ）measuring box.

\subsection{(3 廾) 14 算}

算（サン）calculation.

算出（サンシュツ）calculation; computation.

加算（カサン）addition.

引き算（ひきザン）subtraction.

公算（コウサン）probability; likelihood.

\subsection{(4 升) 8 昇}

昇（ショウ）rise up.

\subsection{(4 开) 6 刑 9 型 12 開}

刑（ケイ）(n,n-suf) penalty; sentence; punishment

型（ケイ、かた）model.

血液型（ケツエキがた）blood type.「Ａ型」type a.

開く（ひらく）to open (door, business, eye, mouth, ...)

\section{3 彡 4 手毛加}

彡 is hair; Kangxi radical 59.

手（て）hand.

手がける（てがける）(v1) to make; to do; to produce; to work on.

手洗い（てあらい）restroom; lavatory; toilet; a place for washing hands

毛（モウ、け）hair.

加（カ）add.

\subsection{(3 彡) 7 形 8 参}

形（ケイ、ギョウ、かたち）shape; form.

参（サン）participate.

参加者（サンカシャ）participant; entrant.

参考（サンコウ）reference; consultation.

\section{3 习 6 羽 7 卵 9 飛 10 弱 11 習}

习 is simplified from 習.

羽（ウ、は、はね）feather

卵（ラン、たまご）egg.

飛（ヒ）fly; jump.

飛ぶ（とぶ）to jump.

弱（ジャク）weak.

弱い（よわい）weak.

練習（レンシュウ）training; practice.

習う（ならう）(vt) to learn.

Note that the bottom component of 習 is 自, not 日.

\section{3 土 9 垢 12 塔}

垢（あか）dirt; filth; grime

塔（トウ）tower.
管制塔（カンセイトウ）control tower.

\section{3 夕 8 夜}

夕（ゆう）evening

夜（ヤ、よる）night.

\section{3 彳 9 律後 15 衝}

律（リツ、リチ）law.
黄金律（オウゴンリツ）Golden Rule.

後（ゴ、コウ、のち、あと）after.
後ろ（うしろ）back; behind; rear.
食後（ショクゴ）after a meal; postprandial.
今後（コンゴ）from now on; hereafter.

衝（ショウ）collide.

\section{3 士女 4 氏 5 仕民}

士（さむらい）gentleman; samurai.

士（シ）(suffix)
a qualified person;
a person with a qualified profession.

学士（ガクシ）university graduate (who has obtained a degree).

工学士（コウガクシ）Bachelor of Engineering.

女（おんな）woman

氏 depicts a man bowing to the left.

氏（シ）(suffix,honorific) Mr.; Mrs.. family. clan.

氏（うじ）family name; birth; lineage.

仕（シ）serve.

仕事（シごと）(n) work; job; business; occupation; employment.

仕方（シかた）way; method; manner.

仕方ない。It can't be helped. There's no other way.

仕える（つかえる）(v1,vi) to serve; to work; to attend.

民（ミン、たみ）people

\subsection{8 妾 11 接}

妾 concubine.

接（セツ）touch; contact.

接吻（セップン）kiss.

接ぐ（つぐ）to join (two things into one); to piece together; to graft.

接する（セッする）to come in contact with; to touch.

\subsection{11 婚}

婚（コン）marriage.

新婚（シンコン）newlywed.

\section{3 上下 4 出}

上（うえ）up

下（した）down

出 depicts something coming out of an open box.
出る（でる）(v1,vi) to go out; to exit; to leave.
出来る（できる）(v1,vi) to be able to do.
出来上がる（できあがる）(vi) to be finished; to be completed; to be ready

\section{3 干 4 午牛 6 汗}

干（カン、ほ）dry.
干す（ほす）(vt) to dry (to make something dry).

午（ゴ、うま）noon.

牛（ギュウ、うし）
cow; bull; ox; buffalo.
beef.

汗（あせ）(n) sweat.

汗をかく(exp,v5k) to sweat.

汗を流す(exp,v5s) to work hard; to sweat.

\section{3 土 5 圧}

土（ド、つち）
soil; earth; ground; dirt; ground (as opposed to the heavens).

圧（アツ）pressure.

気圧（キアツ）air pressure.

指圧（シアツ）finger pressure massage.

\section{3 山小 5 石 6 虫 11 鳥}

山（サン、やま）mount; mountain.

小 depicts three sand granules.

小（ショウ）small.

小さい（ちいさい）small.

石（セキ、コク、いし）stone.

虫（むし）insect; bug; cricket; moth

鳥（チョウ、とり）bird; chicken; chicken meat.

\subsection{4 少}

少（ショウ、すく、すこ）a few; a little.

少ない（すくない）(adj-i) few; little; scarce; limited; insufficient.

少し（すこし）(adv,n) small quantity; few; a little.

\subsection{8 岩 9 砂}

岩（ガン、いわ）rock.

砂（サ、シャ、すな）sand.

\subsection{8 𧈧 11 強}

強（キョウ、ゴウ）strong.

強い（つよい）(adj-i) strong.

\subsection{9 風 11 嵐}

風（フウ、かぜ）wind; -like.

嵐（あらし）storm.

\subsection{10 島 14 鳴}

島（トウ、しま）island.

鳴（メイ）chirp.

鳴く（なく）to chirp; to sing (bird).

\section{3 工 6 江 9 紅}

工（コウ、ク）craft.

工作（コウサク）work; construction; handicraft.

工学（コウガク）engineering.

大工（ダイク）carpenter.

木工（モッコウ）carpenter.

工（たくみ）(name) Takumi.

江（え）inlet; bay

紅（くれない）deep red; crimson

\section{3 亡才幺纟}

亡（ボウ）death; destruction; perishment; the deceased.

亡くす（なくす）(vt) to lose something (because he/she/it dies).

マンション\ruby{火}{カ}\ruby{災}{サイ}、%
４\ruby{歳}{サイ}\ruby{女}{ジョ}\ruby{児}{ジ}\ruby{死}{シ}\ruby{亡}{ボウ}%
　きょうだい２人\ruby{重}{ジュウ}\ruby{体}{タイ}
Fire in a large apartment: 4-year girl dead, 2 siblings in critical state.

才（サイ）ability; talent.
In China 3 strokes, in Japan 4 strokes.

天才（テンサイ）genius; prodigy.

幺 depicts a tiny/small/short thread.

纟 is the simplified form of 糹
(left form of 6 糸 (thread)).

\section{3 弓 4 引}

弓（キュウ、ゆみ）bow (archery, violin).

引（イン、ひ）pull.
引用（インヨウ）quotation; citation; reference.
引く（ひく）(vi,vt) to pull.

\section{3 寸}

% https://en.wiktionary.org/wiki/%E5%AF%B8#Japanese
寸depicts a position on the forearm
where the pulse can be palpated by compressing the radial artery.
寸（スン）an ancient unit of length, approximately 3 cm.

一寸（ちょっと）just a minute; short time; just a little.

一寸待って下さい。Please wait a moment.

一寸！Hey!

\subsection{5 付 6 守 7 対 11 得}

付ける（つける）(v1,vt) to attach, join, stick, glue, fasten.

口付け（くちづけ）(n) kiss.

口付ける（くちづける）(v1) to kiss.

日付（ひづけ）date.

守（シュ）protect.

守る（まもる）to protect.

対（タイ）versus...

対する（タイする）to face each other.

得る（える）(v1,vt) to get; to acquire; to obtain; to earn; to win; to gain; to secure; to attain.
納得（ナットク）understanding, agreement.
納得するto consent, agree; to understand the reason for something.

\subsection{6 寺 9 待持 10 時特}

寺（ジ、てら）Buddhist temple.

待つ（まつ）(vt,vi) to wait; to wait for; to await.

持つ（もつ）to hold; to carry; to possess

時（ジ、とき）time.

時代（ジダイ）era.

三国時代（サンゴクジダイ）The Three Kingdoms period.

戦国時代（センゴクジダイ）The Warring States period.

特（トク）special.

特に（トクに）particularly; especially.

特技（トクギ）special skill.

特待（トクタイ）special treatment.

特待生（トクタイセイ）scholarship student.

\section{3 丸 4 円 6 色}

丸（ガン、まる）circle

円（エン）yen (Japanese currency).
円（まる）circle.

色（ショク、いろ）color.

色々（いろいろ）(adj-na) various.

\section{3 川 4 卬 5 母 5 生 7 貝 9 頁}

川（かわ）river

卬 is an uncommon kanji depicting a man standing and a man kneeling.

母 depicts a pair of breasts.

母（ボ、はは、かあ）mother.

母（はは）(humble) mother.

お母さん（おかあさん）(honorific) mother.

生 depicts a sprout, something sprouting from the ground.

生（セイ）nature; sex; gender.

学生（ガクセイ）student.

生まれる（うまれる）(v1,vi) to be born.

貝 depicts a cowry (a kind of seashell used as money in ancient China).

貝（かい）shell; shellfish.

頁（おおがい）originally depicts a head,
but comes to mean ``leaf'' (an organ in plants).

\subsection{(3 川) 10 流}

流す（ながす）to flow (liquid)

\subsection{(3 川 9 頁) 12 順}

順（ジュン）obey.

\subsection{(5 母) 7 毎 9 海}

毎（マイ、ごと）every.

毎日（マイニチ）everyday.

毎月（マイゲツ、マイつき）every month.

毎時（マイジ）every hour.

毎回（マイカイ）every time (every time it happens); every occurrence.

毎年（マイネン、マイとし）every year.

海（カイ、うみ）sea; beach.
The original character has 10 strokes.
Shinjitai replaces the two dots in the middle
with one vertical stroke.

\subsection{(5 生) 7 麦 8 性}

麦（バク、むぎ）wheat.

性（セイ）nature; sex; gender.

男性（だんせい）male.

女性（じょせい）female.

\subsection{(5 生) 11 産}

産（サン）give birth.

不動産（フドウサン）real estate.

\subsection{(4 卬) 7 迎 8 昂 10 留 12 貿}

迎（ゲイ）welcome.

昂る（たかぶる）(vi) to get excited; to get worked up

留（リュウ）detain.

留学（リュウガク）studying abroad (usually at university level).

貿（ボウ）trade.

貿易（ボウエキ）trade (foreign)

\subsection{(7 貝 5 𠀐) 12 貴 15 遺}

貴（キ）precious; honorable.

貴方（あなた）you (referring to someone of equal or lower status).

遺（イ、ユイ）bequeath; leave behind.

遺児（イジ）orphan.

遺物（イブツ）relic; memento.

遺品（イヒン）articles left by the deceased.

遺書（イショ）will; testament.

遺体（イタイ）corpse; remains.

遺言（ユイゴン）will; testament.

遺す（のこす）to bequeath; to leave behind; to save; to reserve.

\subsection{(7 貝 2 刀) 9 則 11 側}

則（のり）law; rule; regulation.

法則（ホウソク）law; rule.

側（がわ、かわ）side

側（そば）vicinity; near; beside

\subsection{(7 貝) 10 唄員 13 損}

唄（バイ）songs with samisen.

員（イン）(suffix) member.

工員（コウイン）factory worker.

会社員（カイシャイン）company employee.

損（ソン）loss.

\subsection{(7 貝) 9 負 12 買 15 賞賣}

負（フ）negative; minus.

負う（おう）to bear; to carry on one's back.

負かす（まかす）(vt) to defeat.

負ける（まける）(v1,vi)
to lose; to be defeated.
to succumb; to give in; to surrender; to yield.
to be inferior to.

買（バイ）buy.

買う（かう）to buy; to purchase

賞（ショウ）prize; award

受賞（ジュショウ）winning (a prize).

賣 is the old form of 7 売.

\subsection{(7 貝 2 匕) 11 頃}

頃（ころ）time of year; season.

\subsection{8 毒}

毒（ドク）poison.

毒ガス（ドクガス）poison gas.

\subsection{8 青 11 清情 12 晴}

青（セイ、あお）(n) blue; green.

青い（あおい）(adj-i) blue; green.

青ざめる（あおざめる）(v1,vi) to become pale.

清（セイ）clear.

清い（きよい）clear; pure; noble.

情（ジョウ）feelings; emotion; passion; sympathy.

晴（セイ、はれ）clear.

\subsection{11 責 16 積 17 績}

責（セキ）blame.

積（セキ）accumulate.

積雪（セキセツ）fallen snow;
snow cover;
snow that has fallen onto the ground.

面積（メンセキ）area (mathematics).

績（セキ）exploits.

\section{3 也 5 他 6 地池}

也 means ``also''.

他（タ、ほか）other.

他人（タニン）another person; other people; stranger.

地（チ、ジ）land (that is being used for an activity).

空き地（あきチ）vacant land.

耕地（コウチ）arable land.

団地（ダンチ）multi-unit apartments.

池（チ、いけ）pond

\section{3 女 6 好 7 姉 8 妹委妻 10 娘}

好き（すき）like; love; prefer

姉（シ、あね）older sister.
姉（あね）(humble) older sister.
お姉さん（おねえさん）(honorific) older sister.

妹（マイ）younger sister.
妹（いもうと）(humble) younger sister.
妹さん（いもうとさん）(honorific) younger sister.

委員（イイン）committee member.
委ねる（ゆだねる）(v1,vt) to entrust to.

夫妻（フサイ）married couple; husband and wife

娘（むすめ）daughter

\section{3 口 4 𧘇 6 衣}

口（くち）mouth.

𧘇 is an unknown radical.

衣 depicts a cloth.

衣（ころも）
clothes;
garment. gown;
robe. coating (glaze; batter; icing).

\subsection{5 古 6 吉𠮷 7 告}

古 consists of 十(ten) and 口(mouth, generation).

古（コ、ふる）old.

古い（ふるい）old (not of person); ancient; obsolete.

吉（キチ、キツ）good luck.

𠮷 is an alternative form of 吉.

告（コク）tell.

告げる（つげる）to inform; to tell.

告白（コクハク）confess (usually of love).

\subsection{(5 古) 8 固 12 結}

固（コ）hard.

結（ケツ、むす、ゆ）bind; tie (not the wearable).

結う（ゆう）to do up hair.

結ぶ（むすぶ）to bind; to tie; to link.

結合（ケツゴウ）combination; union; binding; catenation; coupling; joining; bond.

結婚（ケッコン）marriage.

結婚式（ケッコンシキ）marriage ceremony.

\subsection{(5 古) 8 周 11 週 14 滴 15 調}

周（シュウ）circumference.

週（シュウ）week.

滴（しずく）a drop of water; a drip.

滴る（したたる）to drip (fall one drop at a time).

調（チョウ）investigate.

\subsection{6 舌 7 乱 9 活 10 舐 13 話}

舌（ゼツ、した）tongue.

乱（ラン）riot.

活（カツ）active.

活きる（いきる）

舐める（なめる）(v1,vt) to lick

話（ワ、はなし）talk.

会話（カイワ）conversation.

話す（はなす）to talk.

\subsection{(4 𧘇) 8 表}

表（ヒョウ）express.

表示（ヒョウジ）manifestation; demonstration. display. representation.

\subsection{(6 衣) 16 壊}

壊（カイ）demolition.

壊す（こわす）(vt) to demolish.

\subsection{(4 𧘇 6 𠮷) 10 袁 13 猿遠園}

袁（エン、オン）robe; long kimono.

猿（エン、さる）monkey.

遠（エン）far.

遠い（とおい）(adj-i) far (spatial distance).

園（エン、その）park; garden; yard; farm.

動物園（ドウブツエン）zoo; animal park; zoological garden.

幼稚園（ヨウチエン）kindergarten.

\subsection{(7 告) 10 造 14 酷}

造る（つくる）to make; to construct.

木造（モクゾウ）wooden; made of wood.

酷（コク）cruel.

\subsection{7 言 14 罰}

言（ゲン、ゴン）say.

言（こと）saying.

言う（いう）to say.

文言（ブンゲン）wording.

言葉（ことば）word; dialect.

罰（バツ）punishment; penalty.

罰金（バッキン）fine; monetary penalty.

罰する（ばっする）to punish; to penalize.

\subsection{10 訓}

訓（クン）instruction.

教訓（キョウクン）lesson; precept; moral instruction.

\subsection{6 名吸 8 味 11 唾問}

名手（メイシュ）expert.

名前（なまえ）
name; full name.
given name; first name.

有名（ユウメイ）(adj-na) fame; famous.

吸（キュウ、す）suck.

吸う（すう）to suck with mouth

味（あじ）flavor; taste.

美味しい（おいしい）delicious.

唾（つば）spit

問（モン）(suffix, counter) counter for questions.

問う（とう）to ask (a question).

質問（シツモン）question; inquiry; enquiry.

\section{(3 walking radical) 6 巡}

巡る（めぐる）to go around.
お巡り（おまわり）policeman.

\section{7 束 10 速 16 頼}

束（ソク、たば）bundle.

速（ソク）fast.

早速（サッソク）(adv) immediately.

急速（キュウソク）rapid (progress).

速い（はやい）fast.

頼（ライ）trust.

頼む（たのむ）to request. to entrust to. to rely on.

頼 was simplified from 賴.

\section{3 子 6 字 8 学乳}

子 depicts a child with arms visible.

子（シ、こ） child.

男子（ダンシ）young man, boy.

女子（ジョシ）young woman, girl.

太子（タイシ）crown prince.

字（ジ）character, letter (of an alphabet, not the letter that is a document)

赤字（あかジ）deficit; red letter; red Han-character.

漢字（カンジ）Han character.

学（ガク）learning, scholarship, erudition, knowledge.

中学（チュウガク）middle school; junior high school.

大学（ダイガク）university.

学界（ガッカイ）academic world.

科学（カガク）science.

乳（ちち）breasts.
授乳（ジュニュウ）breast-feeding.

\section{3 亍 6 行 11 術 12 街}

行（コウ、ギョウ）go.

行く（いく）(vi) to go.

行く（ゆく）(vi) to go.

行う（おこなう）(vt) to do.

術（ジュツ）art.

術者（ジュツシャ）practitioner (of medicine, art, etc.).

街（ガイ、まち）street.

\section{(3 roof radical) 6 安宅 8 空 11 窓}

安全（アンゼン）safety; security.
安心（アンシン）relief; peace of mind.

安い（やすい）cheap; inexpensive; secure.

宅（タク）home; house; residence.
住宅（ジュウタク）residence; housing; residential building.
自宅（ジタク）one's home.
自宅火災（ジタクカサイ）house fire; home fire (disaster).

宅（いえ）home.

空（そら）sky.
空港（クウコウ）airport.
空く（すく）(vi) to become less crowded; to get empty.
空く（あく）(vi) to be open; to be empty.

窓（ソウ、まど）window.

\section{3 女 13 嫌}

嫌い（きらい）hate.

嫌悪感（ケンオカン）aversion.

\section{3 𠫔}

𠫔 (U+20AD4) depicts forearm.

\subsection{4 云 6 会伝 7 芸}

云 depicts clouds.

云う（いう）to say; to call; to name.

会（カイ、エ、あ）meet.

会う（あう）to meet (face to face).

会社（カイシャ）company; corporation.

会社員（カイシャイン）company employee.

年会（ネンカイ）yearly meeting; annual convention.

社会（シャカイ）society.

会見（カイケン）interview.

伝（デン）transmit.

自伝（ジデン）autobiography.

手伝う（てつだう）to help; to assist; to take part in.

伝言（デンゴン）verbal message.

伝記（デンキ）life story.

伝道（デンドウ）proselytizing; evangelism; missionary work.

伝える（つたえる）to convey; to report; to transmit; to communicate.

芸（ゲイ）art.

芸能界（ゲイノウカイ）world of show business.

芸能人（ゲイノウジン）performer; a talent in show business.

\subsection{6 至 9 室}

至（シ）climax.

至言（シゲン）wise saying.

室（シツ、むろ）room.
