\chapter{Kanji 8}

\section{8 京 10 原 13 源}

京（キョウ、ケイ、みやこ）capital.

京（キョウ）imperial capital.

原（ゲン、はら）meadow; field; plain; prairie; tundra; moor; wilderness.

原（ゲン）original; source; raw; origin.
原文（ゲンブン）original text.
原油（ゲンユ）crude oil.
原料（ゲンリョウ）raw materials.

起源（キゲン）origin; beginning; rise.

\subsection{12 就}

就（シュウ）concerning.

\section{8 亟 12 極}

亟（キョク）fast; quick; sudden; urgently; immediately; extremely.

極（キョク、ゴク）poles.
\(10^{48}\) (since the 17th century).

積極的（セッキョクテキ）assertive; proactive.

\section{8 㑒 9 品 10 馬}

㑒 was simplified from the 13-stroke 僉
meaning ``all, together, unanimous''.

品（しな）article; item; thing; goods; stock.

品（ヒン）quality.

品物（しなもの）goods.

作品（サクヒン）work (book; film; composition; etc.). opus.

日用品（ニチヨウヒン）daily necessities.

一品（イッピン）one item; one article; one course of meal.

上品（ジョウヒン）elegant; refined; polished.

品性（ヒンセイ）character (elegant attitude; elegant behavior).

馬（バ、うま、ま）horse.

\subsection{12 検}

検（ケン）examine.

検索（ケンサク）search.

検索エンジンsearch engine.

検索結果（ケンサクケッカ）search results.

コンピューターを検索Search in computer.

Wikipedia内を検索Search in Wikipedia.

検出（ケンシュツ）(n,vs) detection.

検討（ケントウ）consideration.

\ruby{不}{フ}\ruby{正}{セイ}なソフトウェアを検出しました
The implied entity detected malicious software.

\subsection{14 駆駅}

駆（ク）drive.
This was simplified from 驅.

駆ける（かける）(v1,vi) to run (horse).

駅 is simplified from 23 驛.
駅（エキ）train station.

\subsection{16 操}

操る（あやつる）(vt) to be fluent in (a language)

体操（タイソウ）physical exercise; gymnastics; calisthenics.

\subsection{18 験}

験（ケン）verify.

実験（ジッケン）experiment.

治験（チケン）clinical trial.

受験（ジュケン）taking an examination (such as school and university entrance).

経験（ケイケン）experience.

\ruby{未}{み}経験\ruby{者}{しゃ}inexperienced person.

未経験者\ruby{歓}{かん}\ruby{迎}{げい}inexperienced people are welcome.

\section{8 隹}

隹（ふるとり）depicts a short-tailed bird or an old bird.

\subsection{13 催}

催（サイ）sponsor.

開催（カイサイ）holding a meeting; opening an exhibition.

主催（シュサイ）sponsorship.

\subsection{18 雚 13 勧 15 歓 18 観}

雚 depicts a stork or heron.

勧（カン）persuade.

歓（カン）delight.

歓迎（カンゲイ）welcome; reception.

観（カン）observe.

観光（カンコウ）sightseeing.

観光客（カンコウキャク）tourist.

\subsection{12 雇}

雇（コ）employ.

解雇（カイコ）dismissal; firing; layoff.

雇う（やとう）to employ. to hire; to charter.

\subsection{12 集 13 稚 14 雑}

集（シュウ）gather; meet; congregate; swarm; flock.

集める（あつめる）(v1,vt) to collect; to assemble; to gather.

稚（チ）immature; young.

幼稚（ヨウチ）infancy; childish; infantile.

\subsection{13 準}

準（シュン）level; standard.

水準（スイジュン）water level. level; standard.

雜（ザツ）miscellaneous.

\subsection{15 誰 18 曜}

誰（だれ）who

曜（ヨウ）(weekday name).

日曜日（ニチヨウビ）Sunday.

\subsection{18 難}

難（ナン）difficult.

海難（カイナン）shipwreck.

災難（サイナン）disaster.

多難（タナン）(adj-na,n) full of troubles.

難点（ナンテン）fault; weakness.

難しい（むずかしい）difficult.

有難い（ありがたい）(adj-i) grateful; thankful.

\section{8 直 9 県 10 真}

直（ジキ）soon.

県（ケン）prefecture (an administrative division).

直（チョク）direct; in person; frankness; honesty.

直す（なおす）to heal; to cure.
Chinese 直 has 7 strokes.
Japan adds the lower-left corner stroke.

直（ただち）

真（シン、ま）true.

真意（シンイ）real intention; true motive; true meaning.

真っ黒（まっくろ）pitch black.

真っ先（まっさき）the foremost; the beginning.

\section{8 金}

金has a lot to do with metals.

金（キン、かね）gold; money.

金属（キンゾク）metal.
重金属（ジュウキンゾク）heavy metal.
These are also the chemistry terms.

金色（キンいろ、コンジキ）golden (color)

\subsection{13 鉄 14 銅銀 16 鋼}

鉄（テツ、くろがね）iron (lit. 黒金 black metal).

鉄人（てつジン）iron man; strong man.

鉄道（テツドウ）railroad; railway

銅（ドウ、あかがね）copper (lit. 赤金 red metal)

銀（ギン、しろがね）silver (lit. 白金 white metal)

銀行（ギンコウ）bank.

鋼（コウ、はがね）steel

青銅（セイドウ）bronze

鋼鉄（コウテツ）steel

\subsection{14 銃 19 鏡}

銃（ジュウ）gun; small firearms

鏡（かがみ）mirror.

眼鏡（めがね）eyeglasses.

\section{8 実店例}

実（ジツ）truth; reality.
実（み）fruit.
実る（みのる）to bear fruit; to ripen.

店（テン）(n) store; shop.
ラメン店ramen shop (ramen is a kind of Japanese noodle).

店（みせ）.

例えば（たとえば）for example.
例える（たとえる）(v1,vt)
to compare; to liken; to illustrate.
用例（ヨウレイ）example; illustration.

\section{8 和}

和（ワ）peace; Japan.
平和（ヘイワ）peace.
和む（なごむ）(vi) to be softened; to calm down.
和らげる（やわらげる）(v1,vt) to soften; to moderate; to relieve.

\section{8 画}

画 means picture.
画（カク）counter for kanji strokes.
字画（ジカク）number of strokes in a character.
画家（ガカ）painter.
企画（キカク）plan.

\section{8 非 10 俳 12 悲 13 罪}

非（ヒ）un-; non-; negative; mistake; wrong.

非常（ヒジョウ）extraordinary; unusual

非ず（あらず）(exp) no; never mind.

俳（ハイ）haiku.

俳優（ハイユウ）actor; actress.

悲恋（ヒレン）disappointed love

悲しい（かなしい）sad.

罪（ザイ）sin; guilt.

罪（つみ）sin; crime; fault.

犯罪（ハンザイ）crime.

七つの大罪（ななつのダイザイ）seven deadly sins.

\section{(8 雨) 11 雪 12 雲 13 雷電}

雪（セツ、ゆき）snow

雲（ウン、くも）cloud

雷（かみなり）thunder.

電（デン）lightning.
電光（デンコウ）lightning.
電気（デンキ）electricity (lit. lightning spirit).
電話（デンワ）telephone (lit. lightning talk).
電車（デンシャ）electric train (lit. lightning carriage).
電気自動車（デンキジドウシャ）electric car.

\section{8 東 10 凍}

東（トン、ひがし、あずま）east

凍る（こおる）to freeze.
But the kanji for for ice is 氷（こおり）.
