\chapter{Kanji 4}

\section{4 火水 5 氷永 8 泳}

火（カ、ひ）fire, flame.

火事（カジ）fire (disaster).

大火（タイカ）big fire.

水（スイ、みず）water.

井 depicts a square well.

井（い）well (water reservoir).

氷（こおり）ice

永（エイ、なが）eternity.

永い（ながい）(adj-i) very long (time).

泳ぐ（およぐ）(vi) to swim

\subsection{(4 火) 6 灰 7 災 8 炎}

灰（カイ、はい）ashes.

火災（カサイ）fire (disaster).

災い（わざわい）(n) calamity; catastrophe.

炎（ほのお）flame, blaze.

炎天（エンテン）scorching sun.

\subsection{(4 火) 12 焼 15 熱 16 燃}

焼（ショウ）bake; burn.

焼き鳥（やきとり）grilled chicken meat.

熱い（あつい）(adj)hot (temperature)

燃える（もえる）(v1,vi) to burn; to get fired up.

再燃（サイネン）recurrence; revival; resuscitation;
getting spirited/interested once again.

\section{4 心 5 必}

心 is involved in a lot of feeling-related characters.

心配（シンパイ）(adj-na,n,vs) worry, concern, anxiety.

心配（シンパイ）(n,vs) care, help.

必 is unrelated to 心. They only look similar.

必（ヒツ、かなら）inevitable.

必ず（かならず）(adv) always, invariably, certainly.

必要（ヒツヨウ）(adj-na,n) necessity, need.

\subsection{Thoughts: 7 忘 9 思}

忘（ボウ、わす）forget.

忘れる（わすれる）(v1) to forget.

忘年会（ボウネンカイ）year-end party
(lit. forget-year meeting, a meeting to forget the year).

思（シ）think.
思う（おもう）to think

\subsection{10 息}

息（ソク、いき）breath.

息子（むすこ）son

\subsection{13 感}

感じる（カンじる）(v1) to feel.

感心（カンシン）(adj-na,n,vs) admiration. Well done!

\subsection{Other: 9 急 11 悪}

急（キュウ）urgent, sudden, abrupt.

急ぐ（いそぐ）to hurry.

悪（アク）evil, wickedness.

悪人（アクニン）bad person, villain.

悪い（わるい）bad, poor; evil; unprofitable; at fault.

\subsection{(4 心) 7 応 13 愛 17 優}

応え（こたえ）response; reply; answer; solution.

応える（こたえる）(v1) to respond; to reply; to answer.

愛（アイ）(n) love.

愛す（あいす）(vt) to love.

同性愛（ドウセイアイ）homosexuality; same-sex love.

優しい（やさしい）tender; kind; gentle; affectionate; suave

\ruby{愛}{まな}\ruby{弟}{デ}\ruby{子}{シ}favorite pupil/student.

\section{4 水 16 激}

激（ゲキ）violent.

激しい（はげしい）violent.

激しい雨（はげしいあめ）violent rain.

\section{4 中夬 5 央 6 仲 7 決}

中（チュウ、なか）middle.

中学（チュウガク）middle school; junior high school.

…の中（…のなか）middle of something.

田中（たなか）(name) Tanaka.

中山（なかやま）(name) Nakayama.

中川（なかがわ）(name) Nakagawa.

夬Kangxi radical 37.

央（オウ）center.
中央（チュウオウ）center; central.

仲（なか）relation; relationship.

仲間（なかま）company; fellow; colleague; associate; comrade; partner.

決（ケツ）decide.

決まる（きまる）to be decided; to be settled. to look good in clothes.

\subsection{9 映}

映（エイ）reflect.

\section{4 耂 6 考老}

耂 depicts an old man, a bent-over figure with long hair.

考（コウ）consider.

考える（かんがえる）(v1,vt) to consider; to think about.

老い（おい）old age; old (of person); at a late time in life.

老人（ロウジン）old person.

老若（ロウニャク）old and young; all ages.

\subsection{7 孝 11 教}

孝（コウ）filial piety.

教（キョウ）teach.

教育（キョウイク）training; education.

教会（キョウカイ）church.

教え（おしえ）teaching; doctrine.

教える（おしえる）(v1,vt) to teach.

\subsection{8 者 11 著都 14 緒}

者（シャ）(n,suf) someone of that nature; someone doing that work.

者（もの）(n) person (rarely used without a qualifier).

学者（ガクシャ）scholar.

作者（サクシャ）author.

業者（ギョウシャ）trader; merchant.

研究者（ケンキュウシャ）researcher.

著（チョ）renowned.

著書（チョショ）literary work; book; textbook.

著名人（チョメイジン）celebrity.

都（ト、ツ、みやこ）metropolis.

緒（ショ）thong.

一緒（イッショ）together.

\section{4 王 8 国 9 美皇}

王（オウ）king

国（コク、くに） country.

国王（コクオウ）king.

国内（コクナイ）internal; domestic.

共和国（キョウワコク）republic.

インドネシア共和国Republic of Indonesia.

美（ビ）beauty.

美人（ビジン）beautiful woman.

美しい（うつくしい）beautiful.

美味しい（おいしい）delicious (idiosyncratic reading).

皇（コウ）imperial.

皇居（コウキョ）imperial palace.

\section{4 元之文}

元（ゲン、ガン,もと）origin; beginning; source.
元（もと）origin; source.
元々（もともと）
originally; by nature; from the start; since the beginning.

之（これ）this

文（ブン）sentence (literature).
文字（モジ）letter (of alphabet); character (a Han character)
文書（ブンショ）sentence.
文化（ブンカ）culture.
作文（サクブン）writing.
小学生の作文（ショウガクセイのサクブン）elementary-schoolchild writing.

\section{4 今 6 合}

今（コン、いま）now.
今回（コンカイ）this time; this occasion; this occurrence.

合（ゴウ、あ）fit.
合う（あう）to fit.

\section{4 殳 7 投没 10 殺 11 設 15 撃}

殳 depicts a tool or a weapon.

投げる（なげる）to throw

没（ボツ）drowning

殺す（ころす）to kill.

殺害（サツガイ）murder.

殺人（サツジン）murder.

設ける（もうける）to establish.

撃（ゲキ）beat.

電撃（デンゲキ）electric shock.

衝撃（ショウゲキ）shock; crash; impact.

\section{3 尸 4 戸 5 冊 8 雨}

尸 Kangxi radical 44 (``corpse'').

戸（コ、と、べ）door.

神戸（こうべ）Koube (an area in Japan).

雨（あ＼め）rain

\subsection{5 冊 9 柵}

冊（サツ）counter for books.

冊子（サッシ）booklet.

小冊子（ショウサッシ）booklet; pamphlet.

一冊（イッサツ）one book.

柵（サク）fence.

\subsection{6 尺 11 訳}

尺（シャク）shaku (a unit of length).

訳（ヤク）translate.

訳す（ヤクす）(vt) to translate.

\subsection{7 戻 所 10 涙 14 漏 15 編}

戻る（もどる）(vi) to turn back; to return; to go back;
to come back to a previously visited place

所（ところ）place.

近所（キンジョ）neighborhood.

名所（メイショ）famous place.

涙（なみだ）tear (eyewater)

漏れる（もれる）to leak (liquid)

編（ヘン）editing; compilation.

編集（ヘンシュウ）Edit (menu item in computer user interface).

編む（あむ）(vt)
to knit; to plait; to braid.
to compile (an anthology); to edit.

\section{4 勿 8 易 12 陽場}

勿（ブツ）

易（エキ）easy; simple.

易しい（やさしい）(adj-i) easy; plain; simple.

交易（コウエキ）trade; commerce.

易い（やすい）(adj-i) easy (not difficult).

陽（ヨウ）the yang in yin and yang.

太陽（タイヨウ）sun.

場（ジョウ、ば）place.

場所（ばショ）place.

場合（ばあい）case; situation.

\section{4 日月 9 韋}

日（ひ）sun.

日々（ひび）daily; days; old days.

月（つき）moon.

韋 is Kangxi radical 178 (``tanned leather'').

\subsection{5 白 7 伯 8 拍泊迫}

白（ハク）white.

白い（しろい）white.

伯（ハク）chief.

拍（ハク）clap.

泊（ハク）overnight.

迫（ハク）urge.

\subsection{8 昔 10 借}

昔（セキ、むかし）long ago.

借（シャク）borrow.

\subsection{12 散}

散（サン）scatter.

\subsection{12 間温}

間（カン、ケン）interval; space.

間（あいだ）gap; interval; distance; span; stretch (space or time).

間（ま）space; room; time; pause.

人間（ニンゲン）human being.

世間（セケン）world; society.

温かい（あたたかい）(adj-i) warm (of a tangible object; to the touch)

\subsection{9 昼}

昼（チュウ、ひる）daytime.

\subsection{6 早 9 草 12 朝}

早い（はやい）(adj-i) early.

草（くさ）grass

朝（チョウ、あさ）morning.

今朝（けさ）this morning.

早朝（ソウチョウ）early morning.

\subsection{8 明 9 春星}

明（メイ）bright.
明るい（あかるい）bright.

春（シュン、はる）spring (season)

星（セイ、ほし）star

\subsection{(6 早 9 韋) 8 𠦝卓 13 違 16 衛 18 韓}

𠦝 is an alternative form of 卓.

卓（タク）eminent.

違（イ）differ.

相違（ソウイ）difference; discrepancy; variation.

違う（ちがう）(vi) to differ; to not match the correct answer.

間違う（まちがう）to make a mistake; to be incorrect; to be mistaken.

違反（イハン）violation of law.

衛（エイ）defense.

衛生（エイセイ）satellite.

人工衛星（ジンコウエイセイ）human-made satellite.

韓（カン）Korea.

\section{4 公 5 広払 7 私 8 拡}

公（コウ、おおやけ）public; communal; official; governmental.

公安（コウアン）public safety; public welfare.

広（コウ、ひろ）wide.

広い（ひろい）(adj-i) spacious; vast; wide.

広告（コウコク）advertisement.

払（フツ）pay.

私（シ、わたくし、わたし）I; me.

拡（カク）broaden.

\section{4 幻 5 幼}

幻（まぼろし）phantom; vision; illusion; dream

幼（ヨウ、おさな）infancy.
幼い（おさない）very young; immature. childish.

\section{4 心 7 志快}

志（シ、こころざし）intention

快（カイ）cheerful.
快い（こころよい）cheerful.

\section{4 不 7 否}

不（フ）(prefix) not; bad; poor.

不安（フアン）anxiety; insecurity.

不明（フメイ）unknown; obscure; anonymous; unidentified.

否（イナ、いや）negate.

\section{(4 木) 7 村枚}

村（むら）village

枚（マイ）(counter) sheet; thin flat object.
一枚（イチマイ）one sheet.

\section{(4 戈) 6 伐成 7 戒}

伐（バツ）fell; strike; attack; punish.

成（セイ）become.

作成（サクセイ）(n,vs) writing; creation.

コンピュータプログラムを作成するto write a computer program.

戒（カイ）commandment.

十戒（ジッカイ）
(Buddhist) 10 precepts.
(Christian) 10 commandments.

戒める（いましめる）(vt) to admonish.

\section{(4 god radical) 7 社 10 神}

社（シャ）company; firm; association; shrine

社（やしろ）shrine (usually Shinto).

神（かみ）god; spirit; thunder

\section{4 水 7 沈 10 浮}

沈下（チンカ）sinking; subsidence.

沈む（しずむ）(vi) to sink (descend into liquid).
This was simplified from 18 瀋.

浮かぶ（うかぶ）to float (be supported by liquid)

\section{4 水 9 津}

津（つ）seaport; harbor.

津波（つなみ）(n) tsunami; tidal wave.

\section{4 壬 6 任 13 賃}

壬（ニン、ジン、みずのえ）depicts a carrying pole.

任（ニン）responsibility.

賃（チン）fare.

\section{4 天夫 5 矢失}

天（テン）sky; heaven

夫（フ、おっと）husband

矢（シ、や）arrow

失 depicts something falling from a hand.
失う（うしなう）(vt) to lose; to part with.
失明（シツメイ）loss of eyesight.
失血（シッケツ）loss of blood.

\section{4 止 8 歩}

止 depicts a footprint.

止（シ）stop.

止まる（とまる）(vi) to stop (moving); to come to a halt.

止める（やめる）(v1) to stop (doing something).

止める（とめる）(v1) to stop moving (walking, etc.); to park (a car).

歩（ホ、フ、ブ）walk.

歩く（あるく）to walk.

\subsection{6 企}

企（キ）plan.

\subsection{7 足⻊ 14 踊}

足（ソク、あし）foot.
Kangxi radical 157.

⻊ is the component form of 足 (foot).

踊（ヨウ）jump.

\subsection{9 是 12 提 18 題}

是（ゼ）just so.

提（テイ）present.

提供（テイキョウ）offer.

題（ダイ）topic.

問題（モンダイ）problem.

\subsection{5 正 10 症}

正（セイ、ショウ）correct.

不正（フセイ）(adj-na,n) injustice; illegality; fraud.

不正なソフトウェアmalicious software.

正しい（ただしい）right; correct.

症（ショウ）(n,suf) illness.

\subsection{7 走 8 定}

走（ソウ）run.

走る（はしる）(v5r,vi) to run

定（テイ、ジョウ）fix; determine; establish; settle; decide.

安定（アンテイ）stability; equilibrium.

予定（ヨテイ）(n,vs) plan; arrangement; schedule; program.

定住（テイジュウ）settlement; permanent residence.

定める（さだめる）(v1,vt) to decide; to establish; to determine.

未定（ミテイ）not yet fixed; undecided; pending.

\section{4 斤 7 辛}

\subsection{7 近 8 析}

斤 depicts an axe.

辛 depicts a tool used to mark slaves and criminals;
this sometimes also depicts a tree.

辛い（からい）(adj-i) spicy, salty, harsh, hot, acrid.

辛い（つらい）(adj-i) bitter; painful; heartbreaking; difficult.
Suffix づらい(adj-i) means ``difficult to do''.

読みづらい(adj-i) difficult to read.

書きづらい(adj-i) difficult to write.

読みづらい漢字difficult-to-read Han character.

近い（ちかい）(adj-i) near (spatial distance).

近々（ちかぢか）soon.

近作（キンサク）recent work.

最近（サイキン）most recent; recently; these days; nowadays.

析（セキ）chop.

分析（ブンセキ）analysis.

\subsection{13 新 16 親}

新 depicts cutting tree down with axe.

新（シン）new.

新聞（シンブン）news.

新車（シンシャ）new car.

最新（サイシン）newest.

新しい（あたらしい）(adj-i) new.

親（シン、おや）parent.

両親（リョウシン）both parents.

親友（シニュウ）close friend.

母親（ははおや）mother.
