\chapter{Kanji 7}

\section{7 臣 10 姫 17 覧}

臣（シン、ジン）retainer.
Kangxi radical 131.

姫（ひめ）princess.

覧（ラン）perusal.

回覧（カイラン）circulation.

回覧板（カイランバン）circular notice
(especially those distributed to households within a neighborhood association).

\section{7 旲 10 莫 12 募 13 墓 14 慕暮}

旲 is an uncommon kanji.

莫大（バクダイ）enormous; vast.

募（ボ）recruit.

募集（ボシュウ）recruiting; taking applications.

応募（オウボ）application (registration).

墓（ボ、はか）grave; tomb.

慕（ボ）pining.

慕う（したう）(vt) to yearn for; to love dearly.

Xを兄のように慕っている。
The implied entity loves X as if
X were the implied entity's own older brother.

暮（ボ、くら）livelihood.

暮らす（くらす）to live; to get along.

\section{7 赤 9 変 10 恋}

赤い（あかい）(adj-i) red.

変（ヘン）strange.

変わる（かわる）to change; to transform.

変身（ヘンシン）metamorphosis; transformation.

大変（タイヘン）(adv,adj-na,n)very.

恋（レン、こい）romance; love; tender passion.

恋人（こいびと）lover; sweetheart.

恋文（こいぶみ）love letter.

\section{7 車 11 転 15 輪}

車（くるま）car; vehicle. wheel.

水車（スイシャ）water wheel.

転（テン）revolve; turn around; change.

反転（ハンテン）rolling over; turning around.

転ぶ（ころぶ）(vi) to fall down; to fall over.

輪（リン、わ）ring; circle; hoop; wheel.

指輪（ゆびわ）ring (finger accessory).

車輪（シャリン）car wheel.

五輪（ゴリン）the Olympics.

\subsection{9 軍 10 連 12 運}

軍（グン）army.

連（レン）connection; sequence; chaining.

連休（レンキュウ）consecutive holidays.

常連（ジョウレン）regular customer; regular patron.

連れる（つれる）(v1) to lead or take a person.

関連ニュース（カンレンニュース）related news.

運（ウン、はこ）carry; transport.

運ぶ（はこぶ）to carry; to transport; to move.

自動運転（ジドウウンテン）
automatic operation (machine); automatic driving (vehicle).

\section{7 改}

改める（あらためる）(v1,vt) to revise.

\section{7 我 13 義 20 議}

我（ガ、われ）ego; I; me; oneself.

自我（ジガ）self.

無我（ムガ）selflessness.

義（ギ）in-law.

義兄（ギケイ）elder-brother-in-law; elder stepbrother.

義姉（ギシ）elder-sister-in-law; elder stepsister.

義父（ギフ）father-in-law; foster father; stepfather.

定義（テイギ）definition (of terms).

原義（ゲンギ）original meaning.

同義語（ドウギゴ）synonym.

議（ギ）deliberation; consultation; debate; consideration.

議論（ギロン）argument; discussion; dispute; controversy.

\section{(7 言) 9 信計訂 10 記討 11 訪 12 詞 13 詳 14 語 15 談}

信（シン）faith; trust.

信じる（シンじる）(v1,vt) to believe; to have faith in.

計（ケイ）measure; plan.

計画（ケイカク）plan; project; schedule; scheme; program; programme.

計る（はかる）(vt) to measure.

訂正（テイセイ）correction; revision; amendment

記（キ）record.

記録（キロク）record.

記録するto set a record (such as in sports); to break a record.

記録をもつto hold such record.

記事（キジ）article (writing).

選り抜き記事（よりぬきキジ）selected articles.

新しい記事（あたらしいキジ）new articles.

記す（しるす）to record, to write down.

討（トウ）chastise.

討伐（トウバツ）subjugation; suppression.

討つ（うつ）to shoot at; to attack, defeat, destroy, avenge.

訪問（ホウモン）call; visit.

訪中（ホウチュウ）a visit to China.

訪日（ホウニチ）a visit to Japan.

訪米（ホウベイ）a visit to America.

訪れる（おとずれる）(v1,vt) to visit.

訪ねる（たずねる）to visit.

詞（シ、ことば）part of speech; words; poetry.

歌詞（カシ）song lyrics.

作詞（サクシ）song lyrics.

名詞（メイシ）noun.

詳（ショウ、くわ）detailed.

詳細（ショウサイ）details; particulars.

不詳（フショウ）unknown; unidentified; unspecified.

詳しい（くわしい）detailed; full; accurate.

語（ゴ）language.

日本語（ニホンゴ）Japanese language.

英語（エイゴ）English language.

用語（ヨウゴ）term.

言語（ゲンゴ）(linguistic) language.

プログラミング言語programming language.

語る（かたる）(vt) to talk; to tell; to recite.

談（ダン）discuss.

相談（ソウダン）consultation.

示談（ジダン）out-of-court settlement.

座談会（ザダンカイ）symposium; round-table discussion.

\section{7 攻}

攻（コウ、せめ）aggression.
攻める（せめる）(v1,vt) to attack; to assault; to assail.
攻防（コウボウ）attack and defense.

\section{7 君}

君（きみ）you.
…君（…クン）(suffix) Mr. (junior).
君主（クンシュ）ruler; monarch; sovereign.

\section{7 防}

防（ボウ）defense; resistance.

防ぐ（ふせぐ）to resist; to defend against.

防止（ボウシ）prevention; check.

\section{7 助 9 査}

助（すけ）assistance.
助（ジョ）(pref) help; rescue; assistant.
助ける（たすける）(v1,vt) to help.

査（サ）investigate.
巡査（ジュンサ）policeperson.
主査（シュサ）chief examiner; chief investigator.
査問（サモン）enquiry; hearing.

\section{7 医 8 知}

医（イ）medicine; healing; curing; doctor (medical)

知（チ）know.
知る（しる）to know.
日本人の知らない日本語the Japanese language that the Japanese people don't know

\section{7 身}

身（シン）somebody; person.

自身（ジシン）self.

私自身（わたしジシン）I myself; me myself.

出身（シュッシン）person's origin (town, city, country, etc.).

出身地（シュッシンチ）birthplace.

身長（シンチョウ）height (of body).

\section{7 売 13 続 14 読}

売 is simplified from 15 賣.

賣（バイ）sell.

賣る（うる）to sell.

続（ゾク、つづ）continue.

相続（ソウゾク）succession; inheritance.

存続（ソンゾク）duration; continuance.

続く（つづく）to continue.

読（ドク）read.

読者（ドクシャ）reader.

読む（よむ）to read.

\section{7 呆 9 保}

呆 depicts a child.

呆れる（あきれる）(v1,vi) to be amazed, astonished, astounded.

保する（ホする）to guarantee.

保つ（たもつ）to preserve.

保安（ホアン）peace preservation; security.

\section{7 豕 10 家 11 豚 12 象}

豕（いのこ）(pig radical).

家 has at least 3 meanings, depending on how it is read.

家（カ）-er; -ist; someone who does something.

書家（ショカ）calligrapher.

画家（ガカ）painter.

漫画家（マンガカ）Japanese-comic-book-drawing artist.

活動家（カツドウカ）activist.

研究科（ケンキュウカ）researcher.

作家（サッカ）author; creator; writer; artist.

小説家（ショウセツカ）novelist; fiction writer.

政治家（セイジカ）politician; statesman.

作曲家（サッキョクカ）music composer.

史家（シカ）historian.

家（ケ）family.

中川家（なかがわケ）the Nakagawa family.

田中家（たなかケ）the Tanaka family.

マッカーサー家（マッカーサーケ）the MacArthur family; the MacArthurs.

家（うち）house.

「今夜私の家（うち）に来てください。」Please come to my house tonight.

豚（ぶた）pig.

象（ショウ、ゾウ）elephant.

対象（タイショウ）target; object (of worship, study, etc.).

\section{7 谷 10 浴 11 欲}

谷（コク、たに）valley.

浴（ヨク）bathe.

欲（ヨク）longing.

欲しい（ほしい）(adj-i) want.

...欲しい（ほしい）(aux-adj) I want you to ...

食べて欲しいI want you to eat.

黙って欲しいI want you to shut up.
