\chapter{Kanji 2}

\section{2 人 4 仁化 5 代仕付他 6 仲伝伐 7 作 8 供使 9 保便}

化（カ、ばけ）change.

仁（ジン）kindness.

化（カ）(suffix) -ization, -ification.
グローバル化（グローバルカ）globalization.
化学（カガク）chemistry.
化石（カセキ）fossilization.
分化（ブンカ）specialization.

化ける（ばける）(v1,vi) to take the form of.

代（ダイ、タイ）substitute.

代わる（かわる）(vi) to substitute for.

仕（シ、つか）serve.
仕える（つかえる）(v1,vi) to serve; to work; to attend.
仕事（シごと）work.
仕方（シかた）way; method; manner.
仕方ないit can't be helped; there's no other way.

付ける（つける）(v1,vt) to attach, join, stick, glue, fasten.
一寸（ちょっと）just a minute; short time; just a little.
「一寸待って下さい。」``Please wait a moment.''
「一寸！」``Hey!''.
口付け（くちづけ）(n) kiss.
口付ける（くちづける）(v1) to kiss.
日付（ひづけ）date.

他（タ、ほか）other.
他人（タニン）another person; other people; stranger.

仲間（なかま）friend (how does this differ from 友達（ともだち）?)

伝（デン）transmit.
自伝（ジデン）autobiography.
手伝う（てつだう）to help; to assist; to take part in.
伝言（デンゴン）verbal message.
伝記（デンキ）life story.
伝道（デンドウ）proselytizing; evangelism; missionary work.
伝える（つたえる）to convey; to report; to transmit; to communicate.

伐（バツ）fell; strike; attack; punish.

作（サク）work; harvest.
作る（つくる）to make.

供える（そなえる）(v1,vt) to offer; to sacrifice; to dedicate.
子供（こども）child.

使用（シヨウ）(n) use.
使う（つかう）to use.

保する（ホする）to guarantee.
保つ（たもつ）to preserve.
保安（ホアン）peace preservation; security.

便（ベン、ビン）convenience.
不便（フベン）inconvenience.
大便（ダイベン）feces; excrement; shit.
便所（ベンジョ）lavatory.
便り（たより）news; tidings; information; letter.

\section{2 卜 6 外}

外（ガイ、ゲ、そと、ほか、はず）outside; foreign.

外人（ガイジン）foreigner, foreign person.
外国（ガイコク）foreign country.
外界（ガイカイ）outside world.
海外（カイガイ）foreign; abroad; overseas.

外（そと）outside; exterior. open air.

外（ほか）other (places and things); the rest.

外す（はずす）to unfasten. to remove. to leave. to miss (a target).

\section{2 刀 5 刊 7 利判}

刊行（カンコウ）publication; issue.
月刊（ゲッカン）monthly publication; monthly issue.
夕刊（ユウカン）evening newspaper.

利（リ）
advantage; benefit; profit;
interest (the amount added on top of the principal that is paid back).

判（ハン、わか）judge.
判る（わかる）

\section{2 力 5 功}

功（コウ）achievement

\section{2 又 4 収友}

収（シュウ）obtain; income (monetary).
年収（ネンシュウ）annual income.
収める（おさめる）to obtain.

友（ユウ、とも）friend.

\section{(3 walking person radical) 6 行 9 待}

行（コウ、ギョウ、い、ゆ、おこな）go.
行く（いく）(vi) to go.
行く（ゆく）(vi) to go.
行う（おこなう）(vt) to do.

待つ（まつ）(vt,vi) to wait; to wait for; to await.

\section{(3 roof radical) 6 守安宅 8 空定官 9 室 10 家}


守（シュ）protect.

守る（まもる）to protect.

安全（アンゼン）safety; security.
安心（アンシン）relief; peace of mind.

安い（やすい）cheap; inexpensive; secure.

宅（タク）home; house; residence.
住宅（ジュウタク）residence; housing; residential building.
自宅（ジタク）one's home.
自宅火災（ジタクカサイ）house fire; home fire (disaster).

宅（いえ）home.

空（そら）sky.
空港（クウコウ）airport.
空く（すく）(vi) to become less crowded; to get empty.
空く（あく）(vi) to be open; to be empty.

定（テイ、ジョウ）fix; determine; establish; settle; decide.
安定（アンテイ）stability; equilibrium.
予定（ヨテイ）(n,vs) plan; arrangement; schedule; program.
定住（テイジュウ）settlement; permanent residence.
定める（さだめる）(v1,vt) to decide; to establish; to determine.
未定（ミテイ）not yet fixed; undecided; pending.

官 depicts many rooms in a building.
官（カン）government official.
官界（カンカイ）bureaucracy.
長官（チョウカン）secretary; director; chief; director general.

室（シツ、むろ）room.

家 has at least 3 meanings, depending on how it is read.

家（カ）-er; -ist; someone who does something.
書家（ショカ）calligrapher.
画家（ガカ）painter.
漫画家（マンガカ）Japanese-comic-book-drawing artist.
活動家（カツドウカ）activist.
研究科（ケンキュウカ）researcher.
作家（サッカ）author; creator; writer; artist.
小説家（ショウセツカ）novelist; fiction writer.
政治家（セイジカ）politician; statesman.
作曲家（サッキョクカ）music composer.
史家（シカ）historian.

家（ケ）family.
中川家（なかがわケ）the Nakagawa family.
田中家（たなかケ）the Tanaka family.
マッカーサー家（マッカーサーケ）the MacArthur family; the MacArthurs.

家（うち）house.
「今夜私の家（うち）に来てください。」Please come to my house tonight.

\section{3 土 6 地}

地（チ、ジ）land (that is being used for an activity).
空き地（あきチ）vacant land.
耕地（コウチ）arable land.
団地（ダンチ）multi-unit apartments.

\section{3 女 6 好 7 姉 8 妹委妻 10 娘}

好き（すき）like; love; prefer

姉（シ、あね）older sister.
姉（あね）(humble) older sister.
お姉さん（おねえさん）(honorific) older sister.

妹（マイ）younger sister.
妹（いもうと）(humble) younger sister.
妹さん（いもうとさん）(honorific) younger sister.

委員（イイン）committee member.
委ねる（ゆだねる）(v1,vt) to entrust to.

夫妻（フサイ）married couple; husband and wife

娘（むすめ）daughter

\section{3 口 5 古 6 名吸 7 言告 8 味 11 唾}

古 consists of 十(ten) and 口(mouth, generation).
古（コ、ふる）old.
古い（ふるい）old (not of person); ancient; obsolete.

名前（なまえ）
name; full name.
given name; first name.
名手（メイシュ）expert.

吸（キュウ、す）suck.
吸う（すう）to suck with mouth

言contains口(mouth).
言（ゲン、ゴン）.
言（こと）saying.
言う（いう）to say.
言葉（ことば）word; dialect.

告げる（つげる）to inform; to tell.
告白（コクハク）confess (usually of love).

味（あじ）flavor; taste.
美味しい（おいしい）delicious.

唾（つば）spit

\section{(3 walking radical) 5 込 6 巡 7 近 9 送 10 連 12 道}

込む（こむ）to be crowded.

巡る（めぐる）to go around.
お巡り（おまわり）policeman.

近い（ちかい）(adj-i) near (spatial distance).
近々（ちかぢか）soon.
近作（キンサク）recent work.
最近（サイキン）most recent; recently; these days; nowadays.

送る（おくる）to send.
見送る（みおくる）to see off; to escort (for parting); to farewell.
先送る（さきおくる）to postpone.

連（レン）connection; sequence; chaining.
連休（レンキュウ）consecutive holidays.
常連（ジョウレン）regular customer; regular patron.
連れる（つれる）(v1) to lead or take a person.
関連ニュース（カンレンニュース）related news.

道（ドウ、みち）street; road.
鉄道（テツドウ）railway.

\section{(4 old man radical) 6 考 7 孝 8 者}

考（コウ、かんが）consider.
考える（かんがえる）(v1,vt) to consider; to think about.

孝（コウ）filial piety.

者（シャ）(n,suf) someone of that nature; someone doing that work.
者（もの）(n) person (rarely used without a qualifier).
学者（ガクシャ）scholar.
業者（ギョウシャ）trader; merchant.
研究者（ケンキュウシャ）researcher.
作者（サクシャ）author.

\section{(4 god radical) 7 社 10 神}

社（シャ）company; firm; association; shrine

社（やしろ）shrine (usually Shinto).

神（かみ）god; spirit; thunder

\section{4 日 6 早 8 明易 9 春星 12 晴朝陽}

早い（はやい）(adj-i) early.

明（メイ）bright.
明るい（あかるい）bright.

易（エキ）easy; simple.
易しい（やさしい）(adj-i) easy; plain; simple.
交易（コウエキ）trade; commerce.
易い（やすい）(adj-i) easy (not difficult).

春（シュン、はる）spring (season)

星（セイ、ほし）star

晴（セイ、はれ）clear.

朝（チョウ、あさ）morning.
今朝（けさ）this morning.
早朝（ソウチョウ）early morning.

陽（ヨウ）the yang in yin and yang.
太陽（タイヨウ）sun.

\section{4 公 7 私}

公（コウ、おおやけ）public; communal; official; governmental.
公安（コウアン）public safety; public welfare.

私（シ、わたくし、わたし）I; me

\section{4 王 8 国 9 美皇}

国（コク、くに） country.
国王（コクオウ）king.
国内（コクナイ）internal; domestic.

美 beauty.
美味しい（おいしい）delicious (idiosyncratic reading).
美しい（うつくしい）beautiful.

皇（コウ）imperial.
皇居（コウキョ）imperial palace.

\section{4 凶引}

凶（キョウ）evil; villain; bad luck; disaster.
凶悪（キョウアク）(adj-na) atrocious; fiendish; brutal; villainous.

引（イン、ひ）pull.
引用（インヨウ）quotation; citation; reference.
引く（ひく）(vi,vt) to pull.

\section{4 幻 5 幼}

幻（まぼろし）phantom; vision; illusion; dream

幼（ヨウ、おさな）infancy.
幼い（おさない）very young; immature. childish.

\section{4 手 5 打 7 抜 8 抽 11 探}

打（ダ）strike; hit; knock; pound.
安打（アンダ）safe hit (baseball).
打つ（うつ）to hit; to strike; to knock; to beat; to punch; to slap.

抜（バツ、ぬ）slip out.
抜ける（ぬける）(v1,vi) to escape.

抽（チュウ）pluck.
抽出（チュウシュツ）selection (from a group); sampling.

探 depicts a hand groping in a deep cave.
手探り（てさぐり）groping; fumbling.
探す（さがす）(vt)
to search for something lost.
to search for something desired.
探る（さぐる）to feel around for; to fumble for; to grope for.

\section{4 水 Water}

\subsection{Feelings: 7 冷 12 温}

These adjectives describe the feeling when
touching something, not of weather or wind.

冷たい（つめたい）(adj-i) cold (of a tangible object)

温かい（あたたかい）(adj-i) warm (of a tangible object)

\subsection{State: 7 汽 9 洪}

汽（キ）steam.
汽車（キシャ）steam train.

洪水（コウズイ）(n) flood (of liquid)

大水（おおみず）(n) flood (of liquid)

\subsection{Places: 6 江池 8 沼 9 海津}

江（え）inlet; bay

池（チ、いけ）pond

沼（ぬま）swamp; bog

海（カイ、うみ）sea; beach.
The original character has 10 strokes.
Shinjitai replaces the two dots in the middle
with one vertical stroke.

津（つ）seaport; harbor.
津波（つなみ）(n) tsunami; tidal wave.

\subsection{Bodily fluids: 6 汗 10 涙}

汗（あせ）(n) sweat.
汗をかく(exp,v5k) to sweat.
汗を流す(exp,v5s) to work hard; to sweat.

涙（なみだ）tear (eyewater)

\subsection{Action done by the liquid: 8 波 10 凍流}

波（なみ）(n) wave (of liquid)

凍る（こおる）to freeze.
But the kanji for for ice is 氷（こおり）.

流す（ながす）to flow (liquid)

\subsection{Action done to or with the liquid: 7 没沈 8 泳 9 洗 10 浮}

没（ボツ）drowning

沈下（チンカ）sinking; subsidence.
沈む（しずむ）(vi) to sink (descend into liquid).
This was simplified from 18 瀋.

泳ぐ（およぐ）(vi) to swim

洗う（あらう）(vt) to wash

浮かぶ（うかぶ）to float (be supported by liquid)

\subsection{Drinks: 10 酒}

酒（さけ）sake (a Japanese liquor)

\section{5 冬 10 夏}

冬（トウ、ふゆ）winter

夏（カ、なつ）summer

\section{5 圧広 7 応}

圧（アツ）pressure.
気圧（キアツ）air pressure.
指圧（シアツ）finger pressure massage.

広（コウ、ひろ）wide.
広い（ひろい）(adj-i) spacious; vast; wide.
広告（コウコク）advertisement.

応え（こたえ）response; reply; answer; solution.
応える（こたえる）(v1) to respond; to reply; to answer.

\section{5 可}

可（カ）passable; acceptable; tolerable.
許可（キョカ）permission; authorization; approval.

\section{5 犯 8 狙 9 狭}

犯（ハン、おか）crime.
犯す（おかす）to commit (a crime); to perpetrate (a crime).

狙う（ねらう）(vt) to aim at

狭める（せばめる）(v1,vt) to narrow.
狭い（せまい）(adj-i) narrow; confined; small.

\section{5 写}

写生（シャセイ）sketch.
写真（シャシン）multimedia; photograph; movie
写す（うつす）(vt) to photograph.

\section{5 目 7 見 9 相省冒 11 現眼 12 覚}

見る（みる）(v1,vt) to see.
見える（みえる）(v1,vi) to appear.

相（ソウ、あい）mutual.
相手（あいて）companion; partner; company.

相（ショウ）minister.
首相（シュショウ）prime minister; chancellor; premier.

省（ショウ）(suf) ministry; department.
国交省（コッコウショウ）(abbr)
Ministry of Land, Infrastructure, Transport, and Tourism.
省く（はぶく）(vt)
to omit; to leave out; to exclude.
to curtail; to save; to cut down; to economize.

冒 depicts a hat obstructing the sight, implying rashness
(acting without enough thought).
冒す（おかす）(vt) to risk.
冒険（ボウケン）adventure.

現す（あらわす）(vt) to reveal; to show; to display.
現れる（あらわれる）(v1,vi) to appear; to become visible; to materialize.

眼（まなこ）eyeball.
両眼（リョウガン）both eyes.

覚める（さめる）(v1,vi) to wake up

\section{6 舌 9 活}

活（カツ）active.

活きる（いきる）

\section{6 寺}

寺（ジ、てら）Buddhist temple.

\section{6 交 10 校}

交（コウ、まじ）intersect.
交わる（まじわる）cross; intersect; join; meet.

国交（コッコウ）diplomatic relations

学校（ガッコウ）school

\section{6 式}

式（シキ）ceremony.

式（シキ）style.
公式（コウシキ）formal; official.
公式ブログ official blog.

式（シキ）numerical formula.

\section{6 次}

次（つぎ）next (in sequence)

\section{6 先}

先（さき）before; ahead; previous; future.
先生（センセイ）teacher; master; doctor.

\section{6 気}

気（キ）spirit; mind; air; atmosphere
元気（ゲンキ）health(y); vigor; vitality; spirit.
天気（テンキ）weather.
気持ち（きもち）feeling.

\section{7 攻}

攻（コウ、せめ）aggression.
攻める（せめる）(v1,vt) to attack; to assault; to assail.
攻防（コウボウ）attack and defense.

\section{7 君}

君（きみ）you.
…君（…クン）(suffix) Mr. (junior).
君主（クンシュ）ruler; monarch; sovereign.

\section{7 防}

防（ボウ）defense; resistance.
防ぐ（ふせぐ）to resist; to defend against.

\section{7 対}

対（タイ）versus...
対する（タイする）to face each other.

\section{7 走 8 歩}

走（ソウ）.
走る（はしる）(v5r,vi) to run

歩（ホ、フ、ブ）walk.
歩く（あるく）to walk.

\section{7 助 9 査}

助（すけ）assistance.
助（ジョ）(pref) help; rescue; assistant.
助ける（たすける）(v1,vt) to help.

査（サ）investigate.
巡査（ジュンサ）policeperson.
主査（シュサ）chief examiner; chief investigator.
査問（サモン）enquiry; hearing.

\section{4 火 6 灰 7 災 8 炎}

灰（カイ、はい）ashes.

災い（わざわい）(n) calamity; catastrophe.
火災（カサイ）fire (disaster).

炎（ほのお）flame, blaze.
炎天（エンテン）scorching sun.

\section{4 戸 7 戻 8 所}

戻る（もどる）(vi) to turn back; to return; to go back;
to come back to a previously visited place

所（ところ）place.
近所（キンジョ）neighborhood.
名所（メイショ）famous place.

\section{4 心 7 志快}

志（シ、こころざし）intention

快（カイ、こころよ）cheerful.
快い（こころよい）cheerful.

\section{4 不 7 否}

否（イナ、いや）negate.

\section{7 困}

困（コン、こま）quandary.
困る（こまる）(vi) to be troubled; to be embarrassed

\section{7 医}

医（イ）medicine; healing; curing; doctor (medical)

\section{7 身}

身（シン）somebody; person.
自身（ジシン）self.
私自身（わたしジシン）I myself; me myself.
出身（シュッシン）person's origin (town, city, country, etc.).
出身地（シュッシンチ）birthplace.
身長（シンチョウ）height (of body).

\section{4 木 7 村枚}

村（むら）village

枚（マイ）(counter) sheet; thin flat object.
一枚（イチマイ）one sheet.

\section{8 的}

的（テキ）(suffix) -like; typical.
男性的（ダンセイテキ）manly.

的（テキ、まと）mark; target.
目的（モクテキ）purpose; goal; aim; objective; intention.

\section{8 表}

表示（ヒョウジ）manifestation; demonstration. display. representation.

\section{8 学乳}

学（ガク）learning, scholarship, erudition, knowledge.
中学（チュウガク）middle school; junior high school.
大学（ダイガク）university.
学界（ガッカイ）academic world.
科学（カガク）science.

乳（ちち）breasts.
授乳（ジュニュウ）breast-feeding.

\section{8 和}

和（ワ）peace; Japan.
平和（ヘイワ）peace.
和む（なごむ）(vi) to be softened; to calm down.
和らげる（やわらげる）(v1,vt) to soften; to moderate; to relieve.

\section{8 知昂}

知（チ）know.
知る（しる）to know.
日本人の知らない日本語the Japanese language that the Japanese people don't know

昂る（たかぶる）(vi) to get excited; to get worked up

\section{8 性青毒}

性（セイ）nature; sex; gender.
男性（だんせい）male.
女性（じょせい）female.

青（あお）(n) blue; green.
青い（あおい）(adj-i) blue; green.
青ざめる（あおざめる）(v1,vi) to become pale.

毒（ドク）poison.
毒ガス（ドクガス）poison gas.

\section{8 画}

画 means picture.
画（カク）counter for kanji strokes.
字画（ジカク）number of strokes in a character.
画家（ガカ）painter.
企画（キカク）plan.

\section{9 音専負臭変郎飛}

音（オン、おと、ね）sound.

専門家（センモンカ）expert; specialist

負（フ）negative; minus.
負う（おう）to bear; to carry on one's back.
負かす（まかす）(vt) to defeat.
負ける（まける）(v1,vi)
to lose; to be defeated.
to succumb; to give in; to surrender; to yield.
to be inferior to.

臭: In China 10 strokes, in Japan 9 strokes.
The lower character is 犬 in China and 大 in Japan.
臭い（くさい）(adj-i) stinking; malodorous; ill-smelling.

変（ヘン）strange.
変わる（かわる）to change; to transform.
変身（ヘンシン）metamorphosis; transformation.
大変（タイヘン）(adv,adj-na,n)very.

郎（ロウ）son.
野郎（ヤロウ）rascal.

飛（ヒ）fly; jump.
飛ぶ（とぶ）to jump.

\section{9 紅}

紅（くれない）deep red; crimson

\section{9 叙}

叙（ジョ）confer; relate; narrate; describe.
自叙伝（ジジョデン）autobiography.

\section{9 界}

界（カイ）world.
世界（セカイ）world.
学界（ガッカイ）academic world.
業界（ギョウカイ）industry world, business world.

\section{9 胃秋政}

胃（イ）stomach.

秋 depicts the burning of plant stalks (after harvest).
秋（シュウ、あき）autumn; fall season.

政（セイ、まつりごと）rule; government.

\section{7 売 10 員 11 側 12 買 15 賞賣}

売 is simplified from 賣.

員（イン）(suffix) member.
工員（コウイン）factory worker.
会社員（カイシャイン）company employee.

側（がわ、かわ）side

側（そば）vicinity; near; beside

買（バイ）buy.
買う（かう）to buy; to purchase

賞（ショウ）prize; award
受賞（ジュショウ）winning (a prize).

賣（バイ）.
賣る（うる）to sell.

\section{10 原 13 源}

原（ゲン、はら）meadow; field; plain; prairie; tundra; moor; wilderness.

原（ゲン）original; source; raw; origin.
原文（ゲンブン）original text.
原油（ゲンユ）crude oil.
原料（ゲンリョウ）raw materials.

起源（キゲン）origin; beginning; rise.

\section{10 配党息}

配（ハイ）distribute.
配本（ハイホン）distribution of books.
配送センター（ハイソウセンター）distribution center.
配る（くばる）to distribute; to deliver.

党（トウ）political party; faction.
党首（トウシュ）political party leader.

息子（むすこ）son

\section{10 舐}

舐める（なめる）(v1,vt) to lick

\section{10 能笑料}

能（のう）talent; gift; function.

笑う（わらう）to laugh.
笑む（えむ）to smile.

料（リョウ）(suffix) material; charge; rate; fee

\section{10 弱 11 習}

弱（ジャク）weak.
弱い（よわい）weak.

習う（ならう）(vt) to learn.
練習（レンシュウ）training; practice.
Note that the bottom character is 自, not 日.

\section{11 部}

部（ブ）section; department; part.
市部（シブ）urban areas.
学部（ガクブ）a department in a faculty in a university.
東部（トウブ）eastern part.
一部.

\section{11 脱}

脱出（ダッシュツ）escape.
The left component is 肉.

\section{11 動}

動（ドウ）motion.
動く（うごく）(vi) to move.
動画（ドウガ）animation, motion picture.
自動車（ジドウシャ）automobile.
動力（ドウリョク）power; motive power.

\section{11 得}

得る（える）(v1,vt) to get; to acquire; to obtain; to earn; to win; to gain; to secure; to attain.
納得（ナットク）understanding, agreement.
納得するto consent, agree; to understand the reason for something.

\section{11 雪 12 雲 13 雷電}

雪（セツ、ゆき）snow

雲（ウン、くも）cloud

雷（かみなり）thunder.

電（デン）lightning.
電光（デンコウ）lightning.
電気（デンキ）electricity (lit. lightning spirit).
電話（デンワ）telephone (lit. lightning talk).
電車（デンシャ）electric train (lit. lightning carriage).
電気自動車（デンキジドウシャ）electric car.

\section{11 問 12 間開 14 聞関}

問（モン）(suffix, counter) counter for questions.
問う（とう）to ask (a question).
質問（シツモン）question; inquiry; enquiry.
問題（モンダイ）problem.

間（カン、ケン）interval; space.
人間（ニンゲン）human being.
間（あいだ）gap; interval; distance; span; stretch (space or time).
間（ま）space; room; time; pause.

開く（ひらく）to open (door, business, eye, mouth, ...)

聞（ブン、モン）hear.

聞く（きく）to hear.

関（カン、せき）barrier; gate

\section{12 飲}

飲む（のむ）to drink (any liquid, not just liquor)

\section{12 番}

番（バン）number.
番組（バンぐみ）television program.

\section{13 罪}

罪（ザイ）sin; guilt.
罪（つみ）sin; crime; fault.
犯罪（ハンザイ）crime.
七つの大罪（ななつのダイザイ）seven deadly sins.

\section{12 集 13 稚}

集（シュウ）gather; meet; congregate; swarm; flock.
集める（あつめる）(v1,vt) to collect; to assemble; to gather.

稚（チ）immature; young.
幼稚（ヨウチ）infancy; childish; infantile.

\section{14 歌駅}

歌（カ、うた）song.
歌声（うたごえ）singing voice.
歌う（うたう）to sing.

駅 is simplified from 23 驛.
駅（エキ）train station.

\section{Cardinal directions: 5 北 6 西 8 東 9 南}

北（ホク、きた）north

西（セイ、サイ、にし）west

東（トン、ひがし、あずま）east

南（ナン、みなみ）south

They combine as in English.

北西（ホクセイ）northwest

北東（ホクトウ）northeast

東北（トウホク）Touhoku (a prefecture)

南西（ナンセイ）southwest

南東（ナントウ）southeast

\section{Thing: 6 件 8 物事}

件（ケン）matter; case; item

物（ブツ、モツ、もの）thing; object; matter.
物語る（ものがたる）(vt) to tell; to indicate.
書物（ショモツ）books.
食べ物（たべもの）food.

仕事（シごと）(n) work; job; business; occupation; employment.
火事（カジ）fire (as a disaster).
ラメン店で火事fire at a ramen shop.
有事（ユウジ）emergency.
無事（ブジ）safety; peace; quietness.

\section{8 金 13 鉄 14 銅銀 16 鋼}

金has a lot to do with metals.
金（キン、かね）gold; money.

金属（キンゾク）metal.
重金属（ジュウキンゾク）heavy metal.
These are also the chemistry terms.

金色（キンいろ、コンジキ）golden (color)

鉄（テツ、くろがね）iron (lit. 黒金 black metal).
鉄人（てつジン）iron man; strong man.

鉄道（テツドウ）railroad; railway

銅（ドウ、あかがね）copper (lit. 赤金 red metal)

銀（ギン、しろがね）silver (lit. 白金 white metal)

鋼（コウ、はがね）steel

青銅（セイドウ）bronze

鋼鉄（コウテツ）steel

\section{Occupation: 5 仕 8 者}

Difference between 者 and 家（カ）: ???

\section{Heart: 4 心 5 必}

心 is involved in a lot of feeling-related characters.

心配（シンパイ）(adj-na,n,vs) worry, concern, anxiety.
心配（シンパイ）(n,vs) care, help.

必 is unrelated to 心. They only look similar.
必（ヒツ、かなら）inevitable.
必ず（かならず）(adv) always, invariably, certainly.
必要（ヒツヨウ）(adj-na,n) necessity, need.

\subsection{Thoughts: 7 忘 9 思}

忘（ボウ、わす）forget.
忘れる（わすれる）(v1) to forget.
忘年会（ボウネンカイ）year-end party
(lit. forget-year meeting, a meeting to forget the year).

思（シ）think.
思う（おもう）to think

\subsection{Feelings: 12 悲 13 意感}

悲しい（かなしい）sad.
悲恋（ヒレン）disappointed love

意（イ）feelings; thoughts.
意欲（イヨク）motivation; will.
意味合い（イミあい）implication; nuance
小生意気（こなまイキ）cheekiness; impudence.

感じる（カンじる）(v1) to feel.

\subsection{Love: 10 恋}

恋（レン、こい）romance; love; tender passion.
恋人（こいびと）lover; sweetheart.
恋文（こいぶみ）love letter.

\subsection{Other: 9 急 11 悪 14 態}

急（キュウ）urgent, sudden, abrupt.
急ぐ（いそぐ）to hurry.

悪（アク）evil, wickedness.
悪人（アクニン）bad person, villain.
悪い（わるい）bad, poor; evil; unprofitable; at fault.

態（ざま）mess; sorry state; plight; sad sight.
変態（ヘンタイ）sexual perversion.

\section{Hand}

\subsection{Arm or hand movements: 7 投 9 指}

投げる（なげる）to throw

指（ゆび）finger.
指す（さす）(vt) to point.

\subsection{Figurative: 8 押 9 持 10 殺 11 設}

押す（おす）to push; to press; to cram into; to force.
to stamp.
to overwhelm.
押し（おし）(n) push.

持つ（もつ）to hold; to carry; to possess

殺す（ころす）to kill.
殺害（サツガイ）murder.
殺人（サツジン）murder.

設ける（もうける）to establish.

\subsection{Giving and taking: 8 取受 11 授}

取る（とる）(vt) take; fetch; take up.
買い取り（かいとり）purchase; sale. purchase on a non-return policy.

受ける（うける）(v1,vt) to receive

授ける（さずける）(v1,vt) to grant; to award.
授受（ジュジュ）give-and-receive.

\section{Blades}

\subsection{Tangible cutting: 4 切}

切 depicts spoon and sword.
切（セツ、サイ、き）cut.
切る（きる）(v5r) cut.
大切（タイセツ）(adj-na,n) important.
一切（イッサイ）absolutely; (when used with negative) at all.

\subsection{Intangible cutting: 4 分}

分 depicts something separated by a blade.
分（フン、ブン）minute.
１分（イップン）one minute.
１２時３４分（ジュウニジサンジュウヨンプン）12:34 (time).

分（わ）understand.
分かる（わかる）to be understood.

分ける（わける）(v1,vt) to divide; to split; to share; to distribute.

\subsection{Swordtip: 4 方}

方 depicts the tip of a sword.
方（ホウ、かた）direction.
方（かた）(honorific) person.
あの方（あのかた）that person.

\subsection{Dissection: 10 剖}

剖（ボウ）dissection.

\subsection{Law: 6 刑 8 法 9 則 14 罰}

刑（ケイ）(n,n-suf) penalty; sentence; punishment

法（ホウ）law; rule; method; principle.

則（のり）law; rule; regulation.
法則（ホウソク）law; rule.

罰（バツ）punishment; penalty.
罰する（ばっする）to punish; to penalize.
罰金（バッキン）fine; monetary penalty.

\subsection{Separation: 7 別}

別 depicts sword cutting bone.
別な（ベツな）(adj-na) different; separate; another
日付別（ひづけベツ）separate by date.

\subsection{Reduction: 12 減}

減（ゲン）reduction; 10\%減 ten percent reduction.

\subsection{Burglar: 13 賊}

賊（ゾク）burglar; robber.
海賊（カイゾク）pirate; sea robber.

\subsection{War: 13 戦}

戦（いくさ）war.
内戦（ナイセン）civil war.
世界大戦（セカイタイセン）World War.

\section{10 造}

造る（つくる）.
木造（モクゾウ）wooden; made of wood.
