\chapter{Kanji 9}

\section{8 录 9 泉}

录 means ``to carve wood'' or ``to record''.

泉（セン、いずみ）spring (source of water).

\subsection{(8 录) 16 録}

録（ロク）record.

登録（トウロク）registration; entry.

録音（ロクオン）recorded audio; audio recording.

\subsection{(9 泉) 15 線}

線（セン、すじ）line; stripe.
line (telephone line).
line (of a railroad).

打線（ダセン）baseball lineup.

前線（ゼンセン）front line; weather front.

\section{9 音 13 暗意 17 闇}

音（オン、おと、ね）sound.

本音（ホンね）real intention; motive.

暗（アン）dark.

暗い（くらい）dark.

意（イ）feelings; thoughts.

意見（イケン）opinion.

意欲（イヨク）motivation; will.

意図（イト）(n,vs) intention; aim; design.

意味（イミ）meaning; significance.

意味合い（イミあい）implication; nuance

小生意気（こなまイキ）cheekiness; impudence.

闇（やみ）darkness.

(?) 闇で全ては黒く見える。Everything looks black in the dark.
（全て or 皆?)

\section{9 罙 11 深探}

罙 means deep.

深（シン）deep.

深い（ふかい）deep.

深める（ふかめる）(v1,vt) to deepen.

探 depicts a hand groping in a deep cave.

手探り（てさぐり）groping; fumbling.

探す（さがす）(vt)
to search for something lost.
to search for something desired.

探る（さぐる）to feel around for; to fumble for; to grope for.

\section{9 専臭}

専門家（センモンカ）expert; specialist

負（フ）negative; minus.
負う（おう）to bear; to carry on one's back.
負かす（まかす）(vt) to defeat.
負ける（まける）(v1,vi)
to lose; to be defeated.
to succumb; to give in; to surrender; to yield.
to be inferior to.

臭: In China 10 strokes, in Japan 9 strokes.
The lower character is 犬 in China and 大 in Japan.
臭い（くさい）(adj-i) stinking; malodorous; ill-smelling.

\section{7 余 9 叙 10 除}

余（ヨ）leave over.

叙（ジョ）confer; relate; narrate; describe.

自叙伝（ジジョデン）autobiography.

除（ジョ、ジ、のぞ）exclude.

除く（のぞく）to exclude.

削除（サクジョ）elimination; cancellation; deletion.

\section{9 胃政}

胃（イ）stomach.

政（セイ、まつりごと）rule; government.

\section{9 面 11 異}

異（イ）uncommon; different; unusual.
異国（イコク）foreign country.
異性（イセイ）different sex; opposite sex.
異なる（ことなる）to differ; to vary; to disagree.

\section{9 首 12 道}

首（シュ、くび）neck

道（ドウ、みち）street; road.
鉄道（テツドウ）railway.

\section{9 乗}

乗（ジョウ）ride.

乗る（のる）(v1) to get on a public transport vehicle; to aboard; to embark.
