\chapter{Logic}

\section{Counting}

\subsection{Numbers: 1 一 2 二十八七九 3 三千万 4 五六 5 半四 6 両百}

一（イチ、ひと）one

二（ニ、ふた）two

三（サン、み）three

四 four

五 five

六 six

七 seven

八 eight

九 nine

十 ten

半（ハン）half

両（リョウ）both

百（ヒャク）hundred

千（セン）thousand

万（マン）ten thousand

\subsection{Rotation: 6 回}

回 depicts a spiral.
回す（まわす）(vt) to turn; to rotate.
一回（イッカイ）once; one time.
一回目（イッカイめ）first.

\section{Contrast and opposition}

\subsection{Negation: 4 不反 12 無}

不（フ）(prefix) not; bad; poor.
不安（フアン）anxiety; insecurity.
不明（フメイ）unknown; obscure; anonymous; unidentified.

反（ハン）anti-.
反する（ハンする）to oppose; to rebel; to revolt.
反体制（ハンタイセイ）anti-establishment.

無（ム） no, -less, without.
無駄（ムダ）uselessness.
無用（ムヨウ）uselessness.
無敵（ムテキ）invincible, unrivaled (lit. no-enemy).
無茶（ムチャ）absurd, unreasonable (lit. no-tea).
無人（ムジン）unmanned (lit. no-human).
無言（ムゴン）silence (lit. no-say).

\section{Opposite: 6 向}

向 depicts a house and a window.
向かい（むかい）(n) facing; opposite; across the street; other side.
向く（むく）to face; to turn toward.
向ける（むける）(v1,vt) to turn towards.
向こう（むこう）opposite side; other side; opposite direction.
向上（コウジョウ）improvement; advancement; progress.

\subsection{Versus: 7 対}

対（タイ）versus...
対する（タイする）to face each other.

\section{Productive abstract concepts}

\subsection{Turning-into: 4 化}

化（カ）(suffix) -ization, -ification.
グローバル化（グローバルカ）globalization.
化ける（ばける）(v1,vi) to take the form of.
化学（カガク）chemistry.
化石（カセキ）fossilization.
分化（ブンカ）specialization.

\subsection{Self: 6 自}

自（ジ）self.
自ら（みずから）(adv) personally.
自在（ジザイ）freely (at will).
自分（ジブン）self (context? example usage?).

\subsection{Repetition: 6 再}

再（サイ）again, re-

再生（サイセイ）playback; rebirth

再開（サイカイ）reopening

再来（サイライ）return, comeback

\subsection{Same: 6 同}

同じ（おなじ）same.
同性愛（ドウセイアイ）same-sex love.

\subsection{Whole: 6 全}

全 depicts a whole piece of jade.

全（ゼン） whole

全部（ゼンブ）altogether; everything

全く（まったく）(adv) completely, entirely, wholly, totally

\subsection{Every: 7 毎}

毎（マイ）every.
毎日（マイニチ）everyday.
毎月（マイゲツ、マイつき）every month.
毎時（マイジ）every hour.
毎回（マイカイ）every time (every time it happens); every occurrence.
毎年（マイネン、マイとし）every year.

\subsection{Most: 12 最}

最（サイ）most.
最も（もっとも）most.
日本の最も高い山（ニホンのもっともたかいやま）Japan's highest mountain.
世界で最も太い人（セカイでもっともふといひと）The fattest person in the world.
最小（サイショウ）smallest.
最大（サイダイ）biggest.
最初（サイショ）first.
最後（サイゴ）last.
最新（サイシン）newest.
最高（サイコウ）best, highest, tallest.

\section{Foreign: 6 外}

6 外（ガイ）foreign (not from somewhere nearby).
外人（ガイジン）foreigner, foreign person.
外国（ガイコク）foreign country.
外界（ガイカイ）outside world.
海外（カイガイ）foreign; abroad; overseas.

\section{Existence and truth}

\subsection{Existence: 6 在存有}

有 depicts a hand holding a piece of 肉(meat).
有る（ある）to exist.

存じる（ゾンじる）(v1,humble) to think, feel, consider, know.
存在（ソンザイ）existence; being.
共存（キョウゾン）coexistence.
存亡（ソンボウ）life-or-death; existence; destiny.

\subsection{Truth: 10 真}

真（シン）truth; reality

\section{Cause and reason: 5 由 6 因}

由（よし）cause; reason.

因（イン）cause; factor

\section{Geometry}

\subsection{Flat: 5 平}

平 can mean flat, level (not tilted), ordinary, plain, non-special.
平ら（たいら）flatness.
平たい（ひらたい）(adj-i) flat; even; level; simple.
平皿（ひらざら）flat dish.
平安（ヘイアン）peace; tranquility.
平気（ヘイキ）coolness; calmness; composure; unconcern.
平日（ヘイジツ）weekday; ordinary day (non-holiday).
平年（ヘイネン）normal (non-leap) year; normal year (related to harvest; weather).
公平（コウヘイ）fairness; impartiality; justice
水平（スイヘイ）level; horizontally
平文（ヘイブン）plain (non-encrypted) text.
平面（ヘイメン）level (flat and not-tilted) surface.

\subsection{Point: 9 点}

点（テン）point; spot; speck; mark.

\subsection{Shape: 7 形}

形（ケイ、かたち）shape; form.

\subsection{Circle: 3 丸}

丸（まる）circle

\subsection{Intersect: 6 交 10 校}

交わる（まじわる）cross; intersect; join; meet

国交（コッコウ）diplomatic relations

学校（ガッコウ）school

\subsection{Corner: 7 角}

角（カク、かど）corner

角（つの）horn (head protrusion)

\section{Spacetime points and intervals: 5 古 7 近 8 若長 9 前 13 新遠}

古 consists of 十(ten) and 口(mouth, generation).
古い（ふるい）old (not of person); ancient; obsolete.

近い（ちかい）(adj-i) near (spatial distance).
近々（ちかぢか）soon.
近作（キンサク）recent work.
最近（サイキン）most recent; recently; these days; nowadays.

若い（わかい）young; at an early time in life.
若年（ジャクネン）the time when one was young.

長（チョウ）
long (distance or time).
leader.
eldest.
長い（ながい）long (distance); long (time).
長女（チョウジョ）eldest daughter; first-born daughter.
市長（シチョウ）mayor (a government official).
身長（シンチョウ）height (of body).
最長（サイチョウ）longest, tallest.
社長（シャチョウ）company president.

老い（おい）old age; old (of person); at a late time in life.
老人（ロウジン）old person.
老若（ロウニャク）old and young; all ages.

前（まえ）before (time), in front of.
午前（ゴゼン）morning; before noon; a.m. (ante meridien).

新 depicts cutting tree down with axe.
新（シン）new.
新しい（あたらしい）new.
新聞（シンブン）news.
新車（シンシャ）new car.

遠い（とおい）(adj-i) far (spatial distance).
