\chapter{People}

\section{People}

\subsection{Followers and friends: 6 仲共 8 供}

仲間（なかま）friend (how does this differ from 友達（ともだち）?)

共（とも）companion; follower; attendant; retinue

子供（こども）child

\subsection{Returning: 7 戻}

戻る（もどる）(vi) to turn back; to return; to go back;
to come back to a previously visited place

\subsection{Children: 9 保}

保する（ほする）to guarantee

\subsection{Child-rearing: 8 学乳}

学（ガク）learning, scholarship, erudition, knowledge.
中学（チュウガク）middle school; junior high school.
大学（ダイガク）university.
学界（ガッカイ）academic world.
科学（カガク）science.

乳（ちち）breasts.
授乳（ジュニュウ）breast-feeding.

\subsection{Spouses: 4 夫 8 妻}

夫（フ、おっと）husband

夫妻（フサイ）married couple; husband and wife

\subsection{Feelings: 6 好 8 委 13 嫌}

好き（すき）like; love; prefer

委員（イイン）committee member.
委ねる（ゆだねる）(v1,vt) to entrust to.

嫌い（きらい）hate

\subsection{4 氏}

氏 depicts a man bowing to the left.
氏（シ）(suffix,honorific) Mr.; Mrs.. family. clan.
氏（うじ）family name; birth; lineage.

\section{Body parts}

\subsection{Externals: 10 唇 16 頭}

口付け（くちづけ）(n) kiss.
口付ける（くちづける）(v1) to kiss.

唇（くちびる）(n) lips.

頭（あたま）head

\subsection{Eye-related: 7 見 9 冒 11 現 12 覚}

見る（みる）(v1,vt) to see.
見える（みえる）(v1,vi) to appear.

冒 depicts a hat obstructing the sight, implying rashness
(acting without enough thought).
冒す（おかす）(vt) to risk.
冒険（ボウケン）adventure.

現す（あらわす）(vt) to reveal; to show; to display.
現れる（あらわれる）(v1,vi) to appear; to become visible; to materialize.

覚める（さめる）(v1,vi) to wake up

\subsection{Sound-related: 6 曲 7 声 14 聞}

曲（キョク）music.
作曲（サッキョク）musical composition.

声（こえ）voice.
This character was simplified from 17 聲.

聞く（きく）to hear

\subsection{Mouth-related: 8 味 11 問}

味（あじ）flavor; taste.
美味しい（おいしい）delicious.

問（モン）(suffix, counter) counter for questions.
問う（とう）to ask (a question).
質問（シツモン）question; inquiry; enquiry.
問題（モンダイ）problem.

\subsection{Mouth-related: 6 名吸 7 告 10 舐 11 唾}

名前（なまえ）
name; full name.
given name; first name.

吸う（すう）to suck with mouth

告げる（つげる）to inform; to tell.
告白（コクハク）confess (usually of love).

舐める（なめる）(v1,vt) to lick

唾（つば）spit

\subsection{Locomotion: 7 走 8 歩}

走る（はしる）(v5r,vi) to run

歩く（あるく）to walk

\subsection{Internals: 12 筋}

筋肉（キンニク）muscle

\section{Work, craftsmanship, and family}

\subsection{Craftmanship: 3 工 7 作}

工（コウ、ク）craft.
工学（コウガク）engineering.
大工（ダイク）carpenter.
木工（モッコウ）carpenter.
工（たくみ）(name) Takumi.

作（サク）work; harvest.
作る（つくる）to make.

\subsection{Occupation: 5 仕 8 者 10 家}

仕事（シごと）work.
仕方（シかた）way; method; manner.
仕方ないit can't be helped; there's no other way.

者（シャ）(n,suf) someone of that nature; someone doing that work.
者（もの）(n) person (rarely used without a qualifier).
学者（ガクシャ）scholar.
業者（ギョウシャ）trader; merchant.
研究者（ケンキュウシャ）researcher.
作者（サクシャ）author.

家（カ）-er; -ist; someone who does something.
書家（ショカ）calligrapher.
画家（ガカ）painter.
漫画家（マンガカ）Japanese-comic-book-drawing artist.
活動家（カツドウカ）activist.
研究科（ケンキュウカ）researcher.
作家（サッカ）author; creator; writer; artist.
小説家（ショウセツカ）novelist; fiction writer.
政治家（セイジカ）politician; statesman.
作曲家（サッキョクカ）music composer.
史家（シカ）historian.

Difference between 者 and 家（カ）: ???

\subsection{Domicile: 10 家}

家（うち）house.
「今夜私の家（うち）に来てください。」Please come to my house tonight.

\subsection{Family: 10 家}

家（ケ）family.
中川家（なかがわケ）the Nakagawa family.
田中家（たなかケ）the Tanaka family.
マッカーサー家（マッカーサーケ）the MacArthur family; the MacArthurs.

\section{Celestials, skies, gods, and flying: 4 日月 5 白 6 早 8 明}

日（ひ）sun.
日々（ひび）daily; days; old days.

月（つき）moon.

白（ハク）white.
白い（しろい）white.

早い（はやい）(adj-i) early.

明（メイ）bright.
明るい（あかるい）bright.

\subsection{9 星 12 朝}

日付（ひづけ）date.日付別（ひづけベツ）separate by date.(context? usage?)

星（ほし）star

朝（あさ）morning.
今朝（けさ）this morning.
早朝（ソウチョウ）early morning.

\subsection{Skies: 4 天 8 空}

天（テン） sky; heaven

空（そら）sky.
空港（クウコウ）airport.
空く（すく）(vi) to become less crowded; to get empty.
空く（あく）(vi) to be open; to be empty.

\subsection{Gods: 5 申 10 神}

申 depicts a bolt of lightning.
申す（もうす）(humble,vt) to say; to speak.

神（かみ）god; spirit; thunder

\subsection{Atmosphere: 6 気}

気（キ）spirit; mind; air; atmosphere

元気（ゲンキ）

天気（テンキ）weather

気持ち（きもち）feeling

\subsection{Seasons: 5 冬 9 春秋 10 夏}

春（はる）spring (season)

夏（なつ）summer

秋 depicts the burning of plant stalks (after harvest).
秋（あき）autumn; fall season.

冬（ふゆ）winter

\subsection{Atmospheric conditions: 8 雨 11 雪 12 雲 13 雷電}

雨（あめ）rain

雪（ゆき）snow

雲（くも）cloud

雷（かみなり）thunder

電（デン） lightning

電光（デンコウ）lightning

電気（デンキ）electricity (lit. lightning spirit)

電話（デンワ）telephone (lit. lightning talk)

電車（デンシャ）electric train (lit. lightning carriage)

電撃（デンゲキ）electric shock

電気自動車（デンキジドウシャ）electric car

\subsection{Feathers: 6 羽 10 弱 11 習}

6 羽（はね）feather

10 弱

弱い（よわい）weak

11 習 (the bottom character is自not日).

習う（ならう）(vt) to learn

練習（レンシュウ）training; practice

\section{Government}

\subsection{King: 4 王 5 玉生 8 国}

4 王（オウ）king

5 玉（ギョク、たま） ball

生depicts a sprout, something sprouting from the ground.
生（セイ）nature; sex; gender.
学生（ガクセイ）student.
生まれる（うまれる）(v1,vi) to be born.

8 国（コク、くに） country.
国王（コクオウ）king.
国内（コクナイ）internal; domestic.

\subsection{5 主 9 美皇}

5 主（おも）chief; main; principal; important

主人公（シュジンコウ）hero; main character

9 美 beauty

美味しい（おいしい）delicious (idiosyncratic reading)

美しい（うつくしい）beautiful

9 皇（コウ）imperial

皇居（コウキョ）imperial palace

\subsection{8 治}

政治（セイジ）politics.
治める（おさめる）(v1,vt)
to dominate; to rule; to govern; to manage.
to tranquilize; to pacify; to subdue.
to suppress.

治る（なおる）(vi) to heal

治（おさむ）(name) Osamu

仙台（センダイ）(city name) Sendai

\subsection{12 塔}

塔（トウ）tower

管制塔（カンセイトウ）control tower

\section{Strength: 7 助 11 動}

助ける（たすける）(v1,vt) to help

動（ドウ）motion.
動く（うごく）(vi) to move.
動画（ドウガ）animation, motion picture.
自動車（ジドウシャ）automobile.
動力（ドウリョク）power; motive power.

\section{Earth}

\subsection{Utilized land: 6 地}

地（チ）land (that is being used for an activity).
空き地（あきチ）vacant land.
耕地（コウチ）arable land.
団地（ダンチ）multi-unit apartments.

\subsection{Dirt: 9 垢}

垢（あか）dirt; filth; grime

\section{Blades}

\subsection{Tangible cutting: 4 切}

切 (spoon and sword).
切る（きる）(v5r) cut.
切（サイ、セツ）.
一切（イッサイ）absolutely; (when used with negative) at all.
大切（タイセツ）(adj-na,n) important.

\subsection{Intangible cutting: 4 分}

分 depicts something separated by a blade.
分ける（わける）(v1,vt) to divide; to split; to share; to distribute.
１分（イップン）one minute.
１２時３４分（ジュウニジサンジュウヨンプン）12:34 (time).

\subsection{Swordtip: 4 方}

方 depicts the tip of a sword.

方（かた）(honorific) person.
あの方（あのかた）that person.

\subsection{Dissection: 10 剖}

剖（ボウ）dissection.

\subsection{Law: 6 刑 8 法 9 則 14 罰}

刑（ケイ）(n,n-suf) penalty; sentence; punishment

法（ホウ）law; rule; method; principle.

則（のり）law; rule; regulation.
法則（ホウソク）law; rule.

罰（バツ）punishment; penalty.
罰する（ばっする）to punish; to penalize.
罰金（バッキン）fine; monetary penalty.

\subsection{Separation: 7 別}

別 depicts sword cutting bone.
別な（ベツな）(adj-na) different; separate; another

\subsection{Reduction: 12 減}

減（ゲン）reduction; 10\%減 ten percent reduction.

\subsection{Burglar: 13 賊}

賊（ゾク）burglar; robber.
海賊（カイゾク）pirate; sea robber.

\subsection{War, savagery, and violence: 13 戦 15 暴 19 爆}

戦（いくさ）war.
内戦（ナイセン）civil war.
世界大戦（セカイタイセン）World War.

暴 depicts the antler of a buck, representing a savage attack, a violence.
暴動（ボウドウ）insurrection; rebellion; revolt; riot; uprising.
暴風（ボウフウ）storm; windstorm; gale.
暴れる（あばれる）(v1,vi) to rage; to act violently.

爆（バク）burst; explode; bomb.
自爆（ジバク）suicide bombing; self-destruct.
水爆（スイバク）hydrogen bomb.
原爆（ゲンバク）atomic bomb; nuclear bomb.
空爆（クウバク）aerial bombing; air raid.
爆殺（バクサツ）killing by bombing.
爆死（バクシ）death by explosion.

\section{Hand}

\subsection{Arm or hand movements: 7 投 9 指}

投げる（なげる）to throw

指（ゆび）finger.
指す（さす）(vt) to point.

\subsection{Ambiguously figurative: 11 探}

探 depicts a hand groping in a deep cave.
手探り（てさぐり）groping; fumbling.
探す（さがす）(vt)
to search for something lost.
to search for something desired.
探る（さぐる）to feel around for; to fumble for; to grope for.

\subsection{Figurative: 8 押 9 持 10 殺 11 設}

押す（おす）to push; to press; to cram into; to force.
to stamp.
to overwhelm.
押し（おし）(n) push.

持つ（もつ）to hold; to carry; to possess

殺す（ころす）to kill.
殺害（サツガイ）murder.
殺人（サツジン）murder.

設ける（もうける）to establish.

\subsection{Giving and taking: 8 取受 11 授}

取る（とる）(vt) take; fetch; take up.
買い取り（かいとり）purchase; sale. purchase on a non-return policy.

受ける（うける）(v1,vt) to receive

授ける（さずける）(v1,vt) to grant; to award.
授受（ジュジュ）give-and-receive.

\subsection{15 撃}

電撃（デンゲキ）electric shock

\section{Water}

\subsection{Feelings: 7 冷 12 温}

These adjectives describe the feeling when
touching something, not of weather or wind.

冷たい（つめたい）(adj-i) cold (of a tangible object)

温かい（あたたかい）(adj-i) warm (of a tangible object)

\subsection{State: 9 洪}

洪水（コウズイ）(n) flood (of liquid)

大水（おおみず）(n) flood (of liquid)

\subsection{Places: 6 江 8 沼 9 海津}

江（え）inlet; bay

沼（ぬま）swamp; bog

海（うみ）sea; beach.
The original character has 10 strokes.
Shinjitai replaces the two dots in the middle
with one vertical stroke.

津（つ）seaport; harbor.
津波（つなみ）(n) tsunami; tidal wave.

\subsection{Bodily fluids: 6 汗 10 涙}

汗（あせ）(n) sweat.
汗をかく(exp,v5k) to sweat.
汗を流す(exp,v5s) to work hard; to sweat.

涙（なみだ）tear (eyewater)

\subsection{Action done by the liquid: 8 波 10 凍流}

波（なみ）(n) wave (of liquid)

凍る（こおる）to freeze.
But the kanji for for ice is 氷（こおり）.

流す（ながす）to flow (liquid)

\subsection{Action done to or with the liquid: 7 没沈 8 泳 9 洗 10 浮}

没（ボツ）drowning

沈む（しずむ）(vi) to sink (descend into liquid)

泳ぐ（およぐ）(vi) to swim

洗う（あらう）(vt) to wash

浮かぶ（うかぶ）to float (be supported by liquid)

\subsection{Drinks: 10 酒}

酒（さけ）sake (a Japanese liquor)

\subsection{14 滴漏}

滴（しずく）a drop of water; a drip.
滴る（したたる）to drip (fall one drop at a time).

漏れる（もれる）to leak (liquid)

\section{Heart: 4 心 5 必}

心 is involved in a lot of feeling-related characters.
心（シン、こころ）heart.
心配（シンパイ）(adj-na,n,vs) worry, concern, anxiety.
心配（シンパイ）(n,vs) care, help.

必 is unrelated to 心. They only look similar.
必ず（かならず）(adv) always, invariably, certainly.
必要（ヒツヨウ）(adj-na,n) necessity, need.

\subsection{Response: 7 応}

応え（こたえ）response; reply; answer; solution.
応える（こたえる）(v1) to respond; to reply; to answer.

\subsection{Thoughts: 7 忘 9 思}

忘れる（わすれる）(v1) to forget.
忘年会（ボウネンカイ）year-end party
(lit. forget-year meeting, a meeting to forget the year).

思う（おもう）to think

\subsection{Feelings: 12 悲 13 意感}

悲しい（かなしい）sad.
悲恋（ヒレン）disappointed love

意（イ）feelings; thoughts.
意欲（イヨク）motivation; will.
意味合い（イミあい）implication; nuance
小生意気（こなまイキ）cheekiness; impudence.

感じる（カンじる）(v1) to feel.

\subsection{Love: 10 恋 13 愛 17 優}

恋（レン、こい）romance; love; tender passion.
恋人（こいびと）lover; sweetheart.
恋文（こいぶみ）love letter.

愛（アイ）(n) love

優しい（やさしい）tender; kind; gentle; affectionate; suave

\subsection{Other: 9 急 11 悪 14 態}

急（キュウ）urgent, sudden, abrupt.
急ぐ（いそぐ）to hurry.

悪（アク）evil, wickedness.
悪人（アクニン）bad person, villain.
悪い（わるい）bad, poor; evil; unprofitable; at fault.

態（ざま）mess; sorry state; plight; sad sight.
変態（ヘンタイ）sexual perversion.

\section{Money: 10 員 11 側 12 買 15 賞賣(売)}

員（イン）(suffix) member.
工員（コウイン）factory worker.
会社員（カイシャイン）company employee.

側（がわ、かわ）side

側（そば）vicinity; near; beside

買う（かう）to buy; to purchase

賞（ショウ）prize; award

賣る（うる）to sell.
This kanji has been simplified to 7 売る.

\section{Communication: 7 言}

言contains口(mouth).
言（こと）saying.
言う（いう）to say.
言葉（ことば）word; dialect.

\subsection{Faith: 9 信}

信（シン）faith; trust.
信じる（シンじる）(v1,vt) to believe; to have faith in.

\subsection{Schemes: 9 計訂}

計（ケイ）plan.
計画（ケイカク）plan; project; schedule; scheme; program; programme.

訂正（テイセイ）correction; revision; amendment

\subsection{Mouth: 13 話 15 談}

話す（はなす）to talk.

談（ダン）discuss.
相談（ソウダン）consultation.
示談（ジダン）out-of-court settlement.
座談会（ザダンカイ）symposium; round-table discussion.

\subsection{Written: 10 記 14 読語}

記（キ）record.
記す（しるす）to record, to write down.
記録（キロク）record.
記事（キジ）article (writing).
選り抜き記事（よりぬきキジ）selected articles.
新しい記事（あたらしいキジ）new articles.

読む（よむ）to read.

語（ゴ）language.
日本語（ニホンゴ）Japanese language.
英語（エイゴ）English language.

\section{Roofs: 6 安宅 9 室}

安い（やすい）cheap; inexpensive

安全（アンゼン）safety; security

安心（アンシン）relief; peace of mind

住宅（ジュウタク）residence; housing; residential building

自宅（ジタク）one's home

自宅火災（ジタクカサイ）house fire; home fire (disaster)

室（むろ）room.

\section{Roads: 12 道 13 違}

道（みち）street; road.
鉄道（テツドウ）railway.

違う（ちがう）(vi) to differ; to not match the correct answer.

\section{Nourishment: 9 食 12 飲}

食べ物（たべもの）food

食物（ショクもの）food

食べる（たべる）(v1) to eat

飲む（のむ）to drink (any liquid, not just liquor)

\section{Illness: 10 症}

症（ショウ）(n,suf) illness

\section{Arrow, medicine, and knowledge: 5 矢 7 医 8 知}

矢（や）arrow

医（イ）medicine; healing; curing; doctor (medical)

日本人の知らない日本語the Japanese language that the Japanese people don't know

\section{Usage: 8 使}

使用（シヨウ）(n) use.
使う（つかう）to use.

\section{Sprout: 8 青毒性 11 清}

性（セイ）nature; sex; gender

男性（だんせい）male

女性（じょせい）female

青（あお）(n) blue; green.
青い（あおい）(adj-i) blue; green.
青ざめる（あおざめる）(v1,vi) to become pale.

毒（ドク）poison.
毒ガス（ドクガス）poison gas.

清い（きよい）clear; pure; noble

\section{Conjunction: 5 且 7 助 8 狙}

且つ（かつ）and.
且又（かつまた）besides; furthermore; moreover

助（すけ）assistance

助（ジョ）(pref) help; rescue; assistant

狙う（ねらう）(vt) to aim at

\section{Thing: 6 件 8 物事}

件（ケン）matter; case; item

物（ブツ、モツ、もの）thing; object; matter.
物語る（ものがたる）(vt) to tell; to indicate.
書物（ショモツ）books.
食べ物（たべもの）food.

仕事（シごと）(n) work; job; business; occupation; employment.
火事（カジ）fire (as a disaster).
有事（ユウジ）emergency.
無事（ブジ）safety; peace; quietness.

\section{Yin-yang: 12 陽}

陽（ヨウ）the yang in yin and yang

太陽（タイヨウ）sun
