\chapter{Traps}

\section{Confusing characters}

\subsection{Spirit and cloth: 4 礻 5 衤}

\subsection{Arrow and loss: 矢失}

\subsection{3 夂夊 4 攵}

In the Japanese language,
these characters become parts of other characters
instead of being used on their own.

夂 depicts two legs followed by something from behind.

夊 depicts a footprint.

攵 is a variant of 攴 depicting a branch and a hand.

\subsection{Samurai and earth: 3 士土}

士(samurai) has longer upper horizontal stroke.
土(earth) has shorter upper horizontal stroke.

\subsection{Hat, sun, moon, meat, inner}

冃(hat)

日(sun)

月(moon)

\subsection{石 (stone) and 右 (right)}

\subsection{人 (person) and 入 (enter)}

\subsection{王 (king) and 生 (sprout)}

\subsection{業菐美}

13 業 contains 业未.
14 菐 contains 业土人.

\section{Calendar}

These kanji readings for today, yesterday, and tomorrow are irregular.

今日（きょう）today

昨日（きのう）yesterday

明日（あした）tomorrow

Names of weekdays.

日曜日（ニチヨウび）Sunday

月曜日（ゲツヨウび）Monday

火曜日（カヨウび）Tuesday

水曜日（スイヨウび）Wednesday

木曜日（モクヨウび）Thursday

金曜日（キンヨウび）Friday

土曜日（ドヨウび）Saturday

毎日（マイニチ）everyday

Expressions.

また明日（あした）see you again tomorrow; means 'again' and 'tomorrow'

\section{Homophones}

\subsection{おさめる: 4 収 8 治 10 納}

納める（おさめる）(v1,vt)
to dedicate; to make an offering to; to pay (fees) to.
収納（シュウノウ）crop; harvest.
未納（ミノウ）default; failure to pay; overdue payment.

治（おさむ）(name) Osamu.
治る（なおる）(vi) to heal.
治める（おさめる）(v1,vt)
to dominate; to rule; to govern; to manage.
to tranquilize; to pacify; to subdue.
to suppress.
政治（セイジ）politics.

\subsection{うつ:}

\section{つくる: 7 作 10 造}

\section{つかえる}

\section{かわる}

\section{Countries}

日本（ニホン）Japan

中国（チュウゴク）People's Republic of China

英国（エイコク）United Kingdom

米国（ベイコク）United States of America

\section{Personal data}

誕生日（タンジョウビ）birthday; birth date; day of birth.

体重（タイジュウ）body weight

好きなもの（すきなもの）likes

嫌いなもの（きらいなもの）dislikes

\section{Nuances}

\subsection{度 and 回}

What is the difference between 度 and 回?

一度one time.

一回one time.
